---
title: "Statistical Computing"
subtitle: "Fall 2026"
format:
   html: 
    page-layout: full
    toc: false
---

# Schedule

```{r load-libraries, echo = F, message = F, eval = F}
library(tidyverse)
library(knitr)
#library(emo)
```

```{r smart-date-setup, echo = F, message = F, eval = F}
# set these:
firstDay = as.Date("2026-08-24")
firstDayID = "Thu"
lastDay = as.Date("2026-12-04")
lectureDays = c("Tue", "Thu")
labDays = c("Tue")
holidays = as.Date(c("2026-09-07", "2026-10-12", "2026-10-13", "2026-11-25", "2026-11-26", "2026-11-27"))

### render table
allDates = seq(from = firstDay, to = lastDay, by = 'day')
schedule = data.frame(allDates,
                        dayOfWeek = weekdays(allDates, abbreviate=T)) %>%
  filter(dayOfWeek %in% c(lectureDays, labDays))
schedule = schedule %>%
  mutate(lab = ifelse(dayOfWeek %in% labDays, TRUE, FALSE)) %>%
  mutate(Topic = ifelse(allDates %in% holidays, "NO CLASS", "")) %>%
  mutate(Date = format(allDates, format = "%b %d")) %>%
  mutate(Date = paste(dayOfWeek, Date)) %>%
  mutate(Week = rep("", nrow(schedule))) %>%
  #mutate(Prepare = rep("", nrow(schedule))) %>%
  #mutate(AE = rep("", nrow(schedule))) %>%
  mutate(Notes = rep("", nrow(schedule))) %>%
  mutate(Lab = rep("", nrow(schedule))) %>%
  mutate(Project = rep("", nrow(schedule))) %>%
  select(Week, Date, Topic, Lab, Notes, Project)
# to be pasted into console then pasted into markdown:
schedule %>%
  kable(format = "markdown")
```

| Week | Date | Topic | Notes | Lab | Project |
|:-----------|:------------|:-----------|:-----------:|:-----------|:-----------|
| 1 | Tue Aug 25 | [welcome](/slides/welcome.html) & types | [💻](/slides/lec01.html)|[lab 0](/labs/lab0.html)  |  |  |
|  | Thu Aug 27 | control flow, loops and functions | [💻](/slides/lec02.html) |  |  |
| 2 | Tue Sep 01 | NAs, lists, NULL and attributes | [💻](/slides/lec03.html) | [lab 1](/labs/lab1.html) |  |
|  | Thu Sep 03 | object oriented programming |  [💻](/slides/lec04.html) |  |  |
| 3 | Tue Sep 08 | subsetting  |  |  |  |
|  | Thu Sep 10 |  |  |  |  |
| 4 | Tue Sep 15 |  |  |  |  |
|  | Thu Sep 17 |  |  |  |  |
| 5 | Tue Sep 22 |  |  |  |  |
|  | Thu Sep 24 |  |  |  |  |
| 6 | Tue Sep 29 |  |  |  |  |
|  | Thu Oct 01 | exam 1 |  |  |  |
| 7 | Tue Oct 06 |  |  |  |  |
|  | Thu Oct 08 |  |  |  |  |
| 8 | Tue Oct 13 | NO CLASS |  |  |  |
|  | Thu Oct 15 |  |  |  |  |
| 9 | Tue Oct 20 |  |  |  |  |
|  | Thu Oct 22 |  |  |  |  |
| 10 | Tue Oct 27 |  |  |  |  |
|  | Thu Oct 29 |  |  |  |  |
| 11 | Tue Nov 03 |  |  |  |  |
|  | Thu Nov 05 |  |  |  |  |
| 12 | Tue Nov 10 |  |  |  |  |
|  | Thu Nov 12 |  |  |  |  |
| 13 | Tue Nov 17 | exam 2 |  |  |  |
|  | Thu Nov 19 |  |  |  |  |
| 14 | Tue Nov 24 |  |  |  |  |
|  | Thu Nov 26 | NO CLASS |  |  |  |
| 15 | Tue Dec 01 | project presentations |  |  | presentations |
|  | Thu Dec 03 |  |  |  |  |

<!-- | Week | Date | Topic | Notes | Lab | Project | -->

<!-- |:-----------|:-----------|:-----------|:----------:|:-----------|:-----------| -->

<!-- | 1 | Tue Aug 26 | welcome | [💻](/slides/01-welcome.html) | [lab 0](/labs/lab0.html) |  | -->

<!-- |  | Thu Aug 28 | control flow, loops and functions | [💻](/slides/lec02.html) |  |  | -->

<!-- | 2 | Tue Sep 02 | NAs, lists, NULL and attributes | [💻](/slides/lec03.html) | [lab 1](/labs/lab1.html) |  | -->

<!-- |  | Thu Sep 04 | object oriented programming | [💻](/slides/lec04.html) |  |  | -->

<!-- | 3 | Tue Sep 09 | tidy data and dplyr | [💻](/slides/lec05.html) | [lab 2](/labs/lab2.html) |  | -->

<!-- |  | Thu Sep 11 | subsetting | [💻](/slides/lec06.html) |  |  | -->

<!-- | 4 | Tue Sep 16 | ggplot2 | [💻](/slides/lec07.html) | [lab 3](/labs/lab3.html) |  | -->

<!-- |  | Thu Sep 18 | functional programming and purrr | [💻](/slides/lec08.html) |  |  | -->

<!-- | 5 | Tue Sep 23 | regular expressions | [💻](/slides/lec09.html) | [lab 4](/labs/lab4.html) |  | -->

<!-- |  | Thu Sep 25 | review |  |  |  | -->

<!-- | 6 | Tue Sep 30 | exam 1 |  |  |  | -->

<!-- |  | Thu Oct 02 | intro to optimization | [💻](/slides/lec11.html) |  |  | -->

<!-- | 7 | Tue Oct 07 | Newton-Raphson | [💻](/slides/lec12.html) | [lab 5](/labs/lab5.html) |  | -->

<!-- |  | Thu Oct 09 | Newton-Raphson examples |  |  |  | -->

<!-- | 8 | Tue Oct 14 | NO CLASS |  |  |  | -->

<!-- |  | Thu Oct 16 | MM principles | [💻](/slides/lec13.html) |  | [project instructions](/project.html) | -->

<!-- | 9 | Tue Oct 21 | MM example | [💻](/slides/lec14.html) | [lab 6](/labs/lab6.html) |  | -->

<!-- |  | Thu Oct 23 | EM algorithm | [💻](/slides/lec15.html) |  |  | -->

<!-- | 10 | Tue Oct 28 | Monte Carlo integration | [💻](/slides/lec16.html) | project workday |  | -->

<!-- |  | Thu Oct 30 | importance sampling | [💻](/slides/lec17.html) |  |  | -->

<!-- | 11 | Tue Nov 04 | complexity and parallelization | [💻](/slides/lec10.html) |  |  | -->

<!-- |  | Thu Nov 06 | exam 2 |  |  |  | -->

<!-- | 12 | Tue Nov 11 | shiny I | [💻](/slides/lec18.html) | [on GitHub](https://github.com/sta323-fa25) |  | -->

<!-- |  | Thu Nov 13 | shiny II |  |  |  | -->

<!-- | 13 | Tue Nov 18 | SQL + databases | [💻](/slides/lec19.html) | finish lab 7 |  | -->

<!-- |  | Thu Nov 20 | project tips |  |  |  | -->

<!-- | 14 | Tue Nov 25 | APIs | [💻](/slides/lec20.html) | project workday |  | -->

<!-- |  | Thu Nov 27 | NO CLASS |  |  |  | -->

<!-- | 15 | Tue Dec 02 | project presentations |  |  | presentations | -->

<!-- |  | Thu Dec 04 | R packages | [💻](/slides/lec23.html) |  |  | -->

<!-- | Week | Date       | Topic                                  |              Notes              | Lab                      | Exam                                  | Project | -->

<!-- |:----------|:----------|:----------|:---------:|:----------|:----------|:----------| -->

<!-- | 0    | Fri Jan 13 | welcome                                |  [💻](/slides/01-welcome.html)  |                          |                                       |         | -->

<!-- | 1    | Tue Jan 17 | git                                    | [💻](/slides/lab0-welcome.html) | [lab 0](/labs/lab0.html) |                                       |         | -->

<!-- |      | Wed Jan 18 | control flow, loops and functions      |    [💻](/slides/lec02.html)     |                          |                                       |         | -->

<!-- |      | Fri Jan 20 | NAs, lists, NULL and attributes        |    [💻](/slides/lec03.html)     |                          |                                       |         | -->

<!-- | 2    | Tue Jan 24 | if it's a function, fix it             |                                 | [lab 1](/labs/lab1.html) |                                       |         | -->

<!-- |      | Wed Jan 25 | object oriented programming            |    [💻](/slides/lec04.html)     |                          |                                       |         | -->

<!-- |      | Fri Jan 27 | tidy data and dplyr                    |    [💻](/slides/lec05.html)     |                          |                                       |         | -->

<!-- | 3    | Tue Jan 31 | computational masonry                  |                                 | [lab 2](/labs/lab2.html) |                                       |         | -->

<!-- |      | Wed Feb 01 | subsetting                             |    [💻](/slides/lec06.html)     |                          |                                       |         | -->

<!-- |      | Fri Feb 03 | ggplot2                                |    [💻](/slides/lec07.html)     |                          |                                       |         | -->

<!-- | 4    | Tue Feb 07 | dream of plots                         |                                 | [lab 3](/labs/lab3.html) |                                       |         | -->

<!-- |      | Wed Feb 08 | functional programming and purrr       |    [💻](/slides/lec08.html)     |                          |                                       |         | -->

<!-- |      | Fri Feb 10 | regular expressions                    |    [💻](/slides/lec09.html)     |                          |                                       |         | -->

<!-- | 5    | Tue Feb 14 | wrangling the written word             |                                 | [lab 4](/labs/lab4.html) |                                       |         | -->

<!-- |      | Wed Feb 15 | catch-up                               |                                 |                          |                                       |         | -->

<!-- |      | Fri Feb 17 | review / questions / exam              |                                 |                          | [exam 1 release](/exams/exam01.html) |         | -->

<!-- | 6    | Tue Feb 21 | no lab: exam 1                         |                                 |                |   exam 1 due                                    |         | -->

<!-- |      | Wed Feb 22 | complexity and parallelization |  [💻](/slides/lec10.html)                                |                          |                                       |         | -->

<!-- |      | Fri Feb 24 | intro to optimization          |  [💻](/slides/lec11.html)                                |                          |                                       |         | -->

<!-- | 7    | Tue Feb 28 | where's zero?                                       |                                 | [lab 5](/labs/lab5.html)                         |                                       |         | -->

<!-- |      | Wed Mar 01 | Newton-Raphson                  |[💻](/slides/lec12.html)                                  |                          |                                       |         | -->

<!-- |      | Fri Mar 03 | Newton-Raphson examples           |                                 |                          |                                       |         | -->

<!-- | 8    | Tue Mar 07 |  continue lab                                      |                                 |                          |                                       |         | -->

<!-- |      | Wed Mar 08 | MM principles   |   [💻](/slides/lec13.html)                              |                          |                                       |         | -->

<!-- |      | Fri Mar 10 | MM example                        |   [💻](/slides/lec14.html)                              |                          |                                       |         | -->

<!-- | 9    | Tue Mar 14 | NO CLASS: Spring break                 |                                 |                          |                                       |         | -->

<!-- |      | Wed Mar 15 | NO CLASS: Spring break                 |                                 |                          |                                       |         | -->

<!-- |      | Fri Mar 17 | NO CLASS: Spring break                 |                                 |                          |                                       |         | -->

<!-- | 10   | Tue Mar 21 | ascend the likelihood                                      |                                 | [lab 6](/labs/lab6.html)                          |                                       |         | -->

<!-- |      | Wed Mar 22 | EM algorithm                      | [💻](/slides/lec15.html)                                 |                          |                                       |         | -->

<!-- |      | Fri Mar 24 | Monte Carlo integration                          |  [💻](/slides/lec16.html)                                |                          |                                       |         | -->

<!-- | 11   | Tue Mar 28 | continue lab                                       |                                 |                          |                                       |         | -->

<!-- |      | Wed Mar 29 | importance sampling                                 | [💻](/slides/lec17.html)                                 |                          |                                       |         | -->

<!-- |      | Fri Mar 31 | review / questions / exam                          |                                 |                          |   [exam 2 release](/exams/exam02.html)                                     |          | -->

<!-- | 12   | Tue Apr 04 |  no lab: exam 2                                       |                               |                          |  exam 2 due                                     |         | -->

<!-- |      | Wed Apr 05 |  shiny I                 |  [💻](/slides/lec18.html)                                |                          |                                       |         | -->

<!-- |      | Fri Apr 07 | shiny II              |                                 |                          |                                       |         | -->

<!-- | 13   | Tue Apr 11 | shiny dashboard                                  |                                 |               [on GitHub](https://github.com/orgs/sta323-sp23/repositories)         |                                  |         | -->

<!-- |      | Wed Apr 12 | SQL                 |                                 |                          |                                       |         | -->

<!-- |      | Fri Apr 14 | project tips                          |                                 |                          |                                       |         | -->

<!-- | 14   | Tue Apr 18 | project workday                                     |                                 |                          |                                       |         | -->

<!-- |      | Wed Apr 19 | APIs                                   |   [💻](/slides/lec20.html)                               |                          |                                       |         | -->

<!-- |      | Fri Apr 21 | covariance & PCA                                    | [💻](/slides/lec21.html)                                |                          |                                       |         | -->

<!-- | 15   | Tue Apr 25 | project presentations              |                                 |                          |                                       |         | -->

<!-- |      | Wed Apr 26 | special topic: cryptanalysis                         |  [💻](/slides/lec22.html)                               |                          |                                       |         | -->
