---
title: "Dimension reduction"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

# Background

## Matrices in R


::: {.columns}
::: {.column width="50%"}
#### Make matrices
```{r}
A <- matrix(c(1, 2, 2, 1), nrow = 2)
A
B <- matrix(c(0, 1, 0, 1), nrow = 2)
B
```

:::
::: {.column width="50%"}

#### Multiply a matrix

Matrix multiplication proceeds row by column.

```{r}
A %*% B # multiplication
```
:::
:::


## Matrix vector multiplication

## Identity Matrix

The identity matrix has 1s on its diagonals and 0s in the off-diagonal entries.

```{r}
identity <- diag(1, nrow = 2)
identity

A %*% identity == A
```

## Positive definite

## Covariance matrix

## Visualizing a covariance matrix

## Eigenvalues and Eigenvectors

# Principal Component Analysis

## Intuition 

## In practice

## Example

# Multi-dimensional scaling (MDS)

## Goal of MDS

- The goal of MDS is to represent $q$ objects (data points) in $p$ dimensions.

- Common target dimensions are $p = 2$ or $p = 3$ for visualization purposes.

- MDS requires that for every pair of data points $(i; j) \in (1, \ldots, q; 1, \ldots, q)$, we observe a dissimilarity measure (often distance) $d_{ij} = d_{ji}$. 

- The goal of MDS is to estimate $p \times q$ parameter matrix $\mathbf{X}$ where the $i$th column $\mathbf{x}_i$ is estimated by minimizing the function

$$
f(\mathbf{X}) = \sum_{1 \leq i \leq j \leq q} w_{ij} (d_{ij} - ||\mathbf{x}_i - \mathbf{x}_j||)^2
$$

where $w_{ij} = w_{ji}$ is a nonnegative weight and $||\cdot||$ is an arbitrary norm, often taken to be the Euclidean ($L_2$) norm.

## Properties of the objective

The function

$$
f(\mathbf{X}) = \sum_{1 \leq i \leq j \leq q} w_{ij} (d_{ij} - ||\mathbf{x}_i - \mathbf{x}_j||)^2
$$

is



<!-- # MDS (an MM solution) -->




