---
title: "Simulation in R"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---


```{r setup}
#| include: false
library(tidyverse)
library(palmerpenguins)
library(patchwork)

set.seed(323)
theme_set(theme_minimal(base_size = 18))

penguins_clean <- penguins |>
  drop_na(species, island, bill_length_mm, flipper_length_mm,
          body_mass_g, sex) |>
  mutate(
    large_penguin = body_mass_g >= 4500,
    species = fct_drop(species),
    island = fct_drop(island),
    sex = fct_drop(sex)
  )

group_counts <- penguins_clean |>
  count(species, island, name = "n_penguins")
```

## Today

By the end, you should be able to:

1. Explain why simulation is useful in statistical computing.
2. Describe a variable's **support**: the values it can take.
3. Simulate categorical, binary, count, and continuous data.
4. Use observed summaries to create simulated data that are "like" real data.
5. Use Monte Carlo simulation to approximate quantities.


## Why simulation? 

Statisticians want to make inferences from data. Simulation lets us

1. construct measures of uncertainty/reliability (e.g. confidence intervals via bootstrapping)
2. approximate quantities that are difficult to calculate analytically (e.g. integrals using Monte Carlo)
3. generate samples from complicated distributions (e.g. Markov chain Monte Carlo)
4. construct reference distributions for statistical tests (e.g. sample from the null distribution)
5. study the behavior of statistical methods (e.g. bias, power, type I error)
6. validate models (simulate from a fitted/proposed model and see whether simulated data resemble the observed data)

. . .

and much more!

## A lighter starting point

We do not need to start with formal probability theory.

To simulate data, we need to specify:

1. the **support**: what values can occur, and
2. the **distribution**: how frequently different values occur.

::: {.callout-note}
A data generative model is a rule for generating data with a particular support.
:::


## Random seeds

Simulations use pseudo-random number generators, so the exact simulated values may change each time
we run the code.

```{r}
set.seed(323)
rnorm(5)
```

. . .

A **random seed** makes the random results reproducible.

```{r}
set.seed(323)
rnorm(5)
```

::: {.callout-note}
Set the seed once near the beginning of an analysis, not before every line of
simulation code.
:::


## Our running example

The `palmerpenguins` data contain measurements on penguins observed near
Palmer Station, Antarctica.

The version we'll work with:

```{r}
#| eval: false
#| label: subset-penguin
library(tidyverse)
library(palmerpenguins)

set.seed(323)
theme_set(theme_minimal(base_size = 18))

penguins_clean <- penguins |>
  drop_na(species, island, bill_length_mm, flipper_length_mm,
          body_mass_g, sex) |>
  mutate(
    large_penguin = body_mass_g >= 4500,
    species = fct_drop(species),
    island = fct_drop(island),
    sex = fct_drop(sex)
  )

group_counts <- penguins_clean |>
  count(species, island, name = "n_penguins")
```

## Look at the data

```{r}
penguins_clean |>
  glimpse()
```

Question for today:

> Can we use simulation to create artificial penguin data that look like the
> real data?

## Support in the penguin data

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
p1 <- penguins_clean |>
  ggplot(aes(x = species)) +
  geom_bar() +
  labs(x = "Species", y = "Count")

p2 <- penguins_clean |>
  mutate(large_penguin = if_else(large_penguin, "yes", "no")) |>
  ggplot(aes(x = large_penguin)) +
  geom_bar() +
  labs(x = "Mass at least 4500g?", y = "Count")

p3 <- group_counts |>
  ggplot(aes(x = n_penguins)) +
  geom_histogram(binwidth = 10, boundary = 0, color = "white") +
  labs(x = "Penguins per species-island group", y = "Count")

p4 <- penguins_clean |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, color = "white") +
  labs(x = "Body mass (g)", y = "Count")

(p1 | p2) / (p3 | p4)
```

## code

```{r}
#| eval: false
p1 <- penguins_clean |>
  ggplot(aes(x = species)) +
  geom_bar() +
  labs(x = "Species", y = "Count")

p2 <- penguins_clean |>
  mutate(large_penguin = if_else(large_penguin, "yes", "no")) |>
  ggplot(aes(x = large_penguin)) +
  geom_bar() +
  labs(x = "Mass at least 4500g?", y = "Count")

p3 <- group_counts |>
  ggplot(aes(x = n_penguins)) +
  geom_histogram(binwidth = 10, boundary = 0, color = "white") +
  labs(x = "Penguins per species-island group", y = "Count")

p4 <- penguins_clean |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, color = "white") +
  labs(x = "Body mass (g)", y = "Count")

(p1 | p2) / (p3 | p4)
```
:::

## Variable type versus distribution

| Variable | Support | One possible data generative model | Intuition |
|---|---|---|---|
| species | categorical | `sample()` | choose among labels |
| large? | binary | `rbinom(size = 1)` | one trial, success/failure |
| # large in 20 | 0, 1, ..., 20 | `rbinom(size = 20)` | 20 independent trials |
| # events | 0, 1, 2, ... | `rpois()` | events occurring at a rate |
| body mass | real numbers | `rnorm()` | values near a center |

::: {.callout-warning}
A variable type is not a distribution. An integer could be binomial, Poisson, or
something else.
:::

## Distributions are data generative modeling choices

`rbinom()` is natural when we count successes from a fixed number of
independent trials with the same probability of success.

Example:

> In a group of 20 penguins, how many have body mass at least 4500g?

. . .

`rpois()` is natural when we count events occurring at some average **rate**.

Example:

> How many penguins are observed in a survey period?

## Simulating categorical data

A categorical variable takes one of several labels.

```{r}
sample(
  x = c("Adelie", "Chinstrap", "Gentoo"),
  size = 10,
  replace = TRUE
)
```

Use `prob =` when some categories should occur more often.

```{r}
sample(c("Adelie", "Chinstrap", "Gentoo"),
       size = 10, replace = TRUE, prob = c(.44, .20, .36))
```

## From real proportions to simulated species
::: {.columns}
::: {.column width="50%"}
```{r}
species_probs <- penguins_clean |>
  count(species) |>
  mutate(prob = n / sum(n))

species_probs
```
:::


::: {.column width="50%"}
```{r}
sim_species <- sample(
  x = species_probs$species,
  size = nrow(penguins_clean),
  replace = TRUE,
  prob = species_probs$prob
)

tibble(species = sim_species) |>
  count(species) |>
  mutate(prop = n / sum(n))
```
:::
:::

## Simulating binomial data

The binomial distribution counts successes in a fixed number of independent
trials with the same probability of success.

```{r}
rbinom(n = 10, size = 1, prob = .30)
```

With `size = 1`, this is binary 0/1 data.

```{r}
rbinom(n = 10, size = 20, prob = .30)
```

With `size = 20`, this is count data with support 0, 1, ..., 20.

## Estimate a binomial probability from data

```{r}
p_large <- mean(penguins_clean$large_penguin)
p_large
```

For a 0/1 variable, the mean is the proportion of 1s.

```{r}
sim_large <- rbinom(
  n = nrow(penguins_clean),
  size = 1,
  prob = p_large
)

mean(sim_large)
```

## A binomial count simulation

Suppose we repeatedly sample groups of 20 penguins.

How many penguins in each group have body mass at least 4500g?

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
binomial_counts <- tibble(
  n_large = rbinom(n = 1000, size = 20, prob = p_large)
)

binomial_counts |>
  ggplot(aes(x = n_large)) +
  geom_histogram(binwidth = 1, boundary = -.5, color = "white") +
  labs(x = "Number of large penguins in a group of 20",
       y = "Count")
```

## code

```{r}
#| eval: false
binomial_counts <- tibble(
  n_large = rbinom(n = 1000, size = 20, prob = p_large)
)

binomial_counts |>
  ggplot(aes(x = n_large)) +
  geom_histogram(binwidth = 1, boundary = -.5, color = "white") +
  labs(x = "Number of large penguins in a group of 20",
       y = "Count")
```
:::

## Simulating counts with a Poisson data generative model

Count variables often have support

\[
0,1,2,3,\ldots
\]

A Poisson data generative model is often used for events occurring at an average **rate**.

```{r}
rpois(n = 20, lambda = 3)
```

Here, `lambda` controls the mean and variance of the count.

## Exercise 1

Show that when the number of trials is large and the success probability is small, a
Poisson distribution can approximate a binomial distribution.


<!-- ## Poisson approximation to the binomial -->

<!-- When `size` is large and `prob` is small, -->

<!-- \[ -->
<!-- \text{Binomial}(n,p) \approx \text{Poisson}(np). -->
<!-- \] -->

<!-- ::: panel-tabset -->
<!-- ## plot -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| message: false -->
<!-- approx_counts <- tibble( -->
<!--   value = c(rbinom(10000, size = 100, prob = .03), -->
<!--             rpois(10000, lambda = 100 * .03)), -->
<!--   distribution = rep(c("binomial", "poisson"), each = 10000) -->
<!-- ) -->

<!-- approx_counts |> -->
<!--   ggplot(aes(x = value)) + -->
<!--   geom_histogram(binwidth = 1, boundary = -.5, color = "white") + -->
<!--   facet_wrap(~ distribution, ncol = 1) + -->
<!--   labs(x = "Count", y = "Frequency") -->
<!-- ``` -->

<!-- ## code -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- approx_counts <- tibble( -->
<!--   value = c(rbinom(10000, size = 100, prob = .03), -->
<!--             rpois(10000, lambda = 100 * .03)), -->
<!--   distribution = rep(c("binomial", "poisson"), each = 10000) -->
<!-- ) -->

<!-- approx_counts |> -->
<!--   ggplot(aes(x = value)) + -->
<!--   geom_histogram(binwidth = 1, boundary = -.5, color = "white") + -->
<!--   facet_wrap(~ distribution, ncol = 1) + -->
<!--   labs(x = "Count", y = "Frequency") -->
<!-- ``` -->
<!-- ::: -->

## A Poisson data generative model

Suppose we count how many penguins are observed in each species-island group.

```{r}
group_counts
```


```{r}
rpois(n = 10, lambda = group_counts$n_penguins)
```

::: {.callout-note}
- The `rpois` (and other data generative functions) are vectorized. 
- Vector recycling occurs when there is a mismatch between the number being simulated and the length of the input parameter vector.
:::

## Simulating continuous data

Examples:

- body mass
- flipper length
- bill length

For unimodal, symmetric distributions, the normal is a good candidate

```{r}
rnorm(n = 10, mean = 4000, sd = 500)
```

## R's four function prefixes

Many distributions in R have four related functions.

| Prefix | Meaning | Example |
|---|---|---|
| `r` | random generation | `rnorm()` |
| `d` | density (continuous) or probability (discrete) | `dnorm()` |
| `p` | cumulative probability | `pnorm()` |
| `q` | quantiles | `qnorm()` |


## What does a density show?

::: columns

::: {.column width="50%"}
For a continuous variable, the density curve shows where values are relatively
common.

- intervals have probability
- exact points do not
- total area under the curve is 1

::: {.callout-note}
The density is high where simulated values are more likely to appear.
:::

:::
::: {.column width="50%"}
::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
tibble(x = seq(2500, 6500, length.out = 200)) |>
  ggplot(aes(x = x)) +
  stat_function(
    fun = dnorm,
    args = list(mean = 4200, sd = 850)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "A normal density curve")
```

## code

```{r}
#| eval: false
tibble(x = seq(2500, 6500, length.out = 200)) |>
  ggplot(aes(x = x)) +
  stat_function(
    fun = dnorm,
    args = list(mean = 4200, sd = 850)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "A normal density curve")
```
:::

:::
:::

## Compare a density to simulated values

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
sim_body <- tibble(
  body_mass_g = rnorm(1000, mean = 4200, sd = 850)
)

sim_body |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 250, color = "white") +
  stat_function(
    fun = dnorm,
    args = list(mean = 4200, sd = 850)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Simulated data and the density used to generate them")
```

## code

```{r}
#| eval: false
sim_body <- tibble(
  body_mass_g = rnorm(1000, mean = 4200, sd = 850)
)

sim_body |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 250, color = "white") +
  stat_function(
    fun = dnorm,
    args = list(mean = 4200, sd = 850)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Simulated data and the density used to generate them")
```
:::


## Is this normal a good fit?

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
body_mean <- mean(penguins_clean$body_mass_g)
body_sd <- sd(penguins_clean$body_mass_g)

penguins_clean |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 250, color = "white") +
  stat_function(
    fun = dnorm,
    args = list(mean = body_mean, sd = body_sd)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Observed body masses and one normal data generative model")
```

## code

```{r}
#| eval: false
body_mean <- mean(penguins_clean$body_mass_g)
body_sd <- sd(penguins_clean$body_mass_g)

penguins_clean |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 250, color = "white") +
  stat_function(
    fun = dnorm,
    args = list(mean = body_mean, sd = body_sd)
  ) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Observed body masses and one normal data generative model")
```
:::


## Simulating data like the data

Suppose we want to create artificial data that are **like** the observed data.

One simple strategy:

1. choose a data generative model,
2. estimate its parameters using summaries of the observed data,
3. simulate new data from that estimated distribution,
4. compare simulated data with observed data.

This is the main idea behind a simple **method of moments** approach.

## Method of moments, informally

A **moment** is a summary of a distribution.

For today:

- match the observed mean
- match the observed variance

## Fit and simulate a normal body-mass data generative model

```{r}
body_moments <- penguins_clean |>
  summarize(
    mean_body = mean(body_mass_g),
    sd_body = sd(body_mass_g),
    n = n()
  )

sim_penguins_1 <- tibble(
  body_mass_g = rnorm(body_moments$n,
                      body_moments$mean_body,
                      body_moments$sd_body)
)

body_moments
```

## Did the simulation work?

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
observed_body <- penguins_clean |>
  transmute(source = "observed", body_mass_g)

sim_body_1 <- sim_penguins_1 |>
  transmute(source = "simulated", body_mass_g)

bind_rows(observed_body, sim_body_1) |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, color = "white") +
  facet_wrap(~ source, ncol = 1) +
  labs(x = "Body mass (g)", y = "Count",
       title = "Observed data versus one simulated data set")
```

## code

```{r}
#| eval: false
observed_body <- penguins_clean |>
  transmute(source = "observed", body_mass_g)

sim_body_1 <- sim_penguins_1 |>
  transmute(source = "simulated", body_mass_g)

bind_rows(observed_body, sim_body_1) |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, color = "white") +
  facet_wrap(~ source, ncol = 1) +
  labs(x = "Body mass (g)", y = "Count",
       title = "Observed data versus one simulated data set")
```
:::

## A problem with the first simulation

Penguin body mass has structure.

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
penguins_clean |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (g)",
       title = "Body mass differs by species")
```

## code

```{r}
#| eval: false
penguins_clean |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (g)",
       title = "Body mass differs by species")
```
:::

One normal curve for all penguins ignores this structure.

## Match and simulate within species

```{r}
species_moments <- penguins_clean |>
  group_by(species) |>
  summarize(n = n(),
            mean_body = mean(body_mass_g),
            sd_body = sd(body_mass_g),
            .groups = "drop")

sim_penguins_2 <- species_moments |>
  rowwise() |>
  mutate(body_mass_g = list(rnorm(n, mean_body, sd_body))) |>
  unnest(body_mass_g) |>
  select(species, body_mass_g)

species_moments
```

## Compare observed and simulated data

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
observed_by_species <- penguins_clean |>
  transmute(source = "observed", species, body_mass_g)

sim_by_species <- sim_penguins_2 |>
  mutate(source = "simulated")

bind_rows(observed_by_species, sim_by_species) |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  facet_wrap(~ source) +
  labs(x = "Species", y = "Body mass (g)",
       title = "A better simulation: match summaries within species")
```

## code

```{r}
#| eval: false
observed_by_species <- penguins_clean |>
  transmute(source = "observed", species, body_mass_g)

sim_by_species <- sim_penguins_2 |>
  mutate(source = "simulated")

bind_rows(observed_by_species, sim_by_species) |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  facet_wrap(~ source) +
  labs(x = "Species", y = "Body mass (g)",
       title = "A better simulation: match summaries within species")
```
:::

## Exercise 2

Use method of moments to simulate flipper lengths within species.

<!-- 1. Compute the observed mean and standard deviation of `flipper_length_mm` -->
<!--    within each species. -->
<!-- 2. Simulate one artificial flipper length for each observed penguin. -->
<!-- 3. Make a plot comparing observed and simulated flipper lengths. -->

## Simulating a whole toy data set

We can combine several pieces:

1. simulate species using observed species proportions,
2. simulate body mass using species-specific normal data generative models,
3. combine the simulated variables into one data frame.

```{r}
sim_penguins_3 <- tibble(
  species = sample(species_probs$species, nrow(penguins_clean),
                   replace = TRUE, prob = species_probs$prob)
) |>
  left_join(species_moments, by = "species") |>
  mutate(body_mass_g = rnorm(n(), mean_body, sd_body)) |>
  select(species, body_mass_g)
```

## Model checking by simulation

::: panel-tabset
## plot

```{r}
#| echo: false
#| message: false
obs_plot <- observed_by_species |>
  mutate(source = "observed")

sim_plot <- sim_penguins_3 |>
  mutate(source = "simulated")

bind_rows(obs_plot, sim_plot) |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = .45) +
  facet_wrap(~ source, ncol = 1) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Does the simulated data look like the observed data?")
```

## code

```{r}
#| eval: false
obs_plot <- observed_by_species |>
  mutate(source = "observed")

sim_plot <- sim_penguins_3 |>
  mutate(source = "simulated")

bind_rows(obs_plot, sim_plot) |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = .45) +
  facet_wrap(~ source, ncol = 1) +
  labs(x = "Body mass (g)", y = "Density",
       title = "Does the simulated data look like the observed data?")
```
:::

## What did we build?

We built a data generative model:

1. choose a species,
2. generate a body mass from a species-specific distribution.

```{r}
#| eval: false
species <- sample(species_values, size = n, replace = TRUE,
                  prob = species_probs)

body_mass <- rnorm(n, mean = species_mean[species],
                   sd = species_sd[species])
```

::: {.callout-note}
Simulation forces us to be explicit about our assumptions.
:::


## Monte Carlo approximation

Monte Carlo approximation means:

> approximate a quantity using many random simulations.

The usual pattern is:

1. simulate many values,
2. calculate something from each simulated value,
3. compute the mean of the results.

This turns difficult mathematical calculations into computing problems.

. . . 

The key idea is that integrals are means, and means are integrals!


# Toy example: estimating $\pi$

## The picture to have in mind

```{r}
#| fig-height: 4
#| echo: false
#| warning: false
#| message: false
#| eval: true
library(ggplot2)
set.seed(2)

# Create a data frame with values for x and y coordinates
df = data.frame(x = cos(seq(0, 2*pi, length.out = 100)), 
                 y = sin(seq(0, 2*pi, length.out = 100)))

df2 = data.frame(x = runif(50, min = -1, max = 1),
                 y = runif(50, min = -1, max = 1))

# Create a ggplot object with the data frame
ggplot(df, aes(x, y)) + 
  geom_path(size = 0.5) +
  coord_equal() +
  geom_rect(aes(xmin = -1, xmax = 1, ymin = -1, ymax = 1), 
                   fill = NA, color = "black") +
  geom_point(data = df2, aes(x = x, y = y)) +
  theme_bw()
```

-   The area of the unit circle, $A_\text{circle} = \pi$. We'll pretend we don't know and want to estimate $A_\text{circle}$

-   50 random points thrown into the box defined by coordinates (-1, -1), (-1, 1), (1, -1), (1, 1)

-   39 points land inside, 11 points outside

-   $\frac{39}{50} \approx \frac{A_\text{circle}}{A_\text{box}}$

-   $A_{\text{circle}} \approx .78 \cdot 4 = 3.12$
