---
title: "Markov chains"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

```{r setup}
#| include: false
library(tidyverse)
library(mvtnorm)

set.seed(323)
theme_set(theme_minimal(base_size = 18))

states <- c("Dry", "Wet")

P_weather <- matrix(
  c(.85, .15,
    .40, .60),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(from = states, to = states)
)

simulate_chain <- function(P, initial, n_steps = 100) {
  states <- rownames(P)

  chain <- character(n_steps + 1)
  chain[1] <- initial

  for (t in seq_len(n_steps)) {
    chain[t + 1] <- sample(
      states,
      size = 1,
      prob = P[chain[t], ]
    )
  }

  chain
}

matrix_power <- function(P, k) {
  out <- diag(nrow(P))

  for (i in seq_len(k)) {
    out <- out %*% P
  }

  dimnames(out) <- dimnames(P)
  out
}

Sigma_geo <- matrix(
  c(1.2, .45,
    .45, .7),
  nrow = 2,
  byrow = TRUE
)

simulate_brownian <- function(n_steps = 100,
                              start = c(0, 0),
                              sigma = diag(2),
                              dt = 1) {

  locations <- matrix(
    NA,
    nrow = n_steps + 1,
    ncol = 2
  )

  locations[1, ] <- start

  for (t in 1:n_steps) {

    move <- rmvnorm(
      1,
      mean = c(0, 0),
      sigma = sigma * dt
    )

    locations[t + 1, ] <- locations[t, ] + move
  }

  tibble(
    step = 0:n_steps,
    east_km = locations[, 1],
    north_km = locations[, 2]
  )
}

nucleotides <- c("A", "C", "G", "T")

jc_P <- function(d) {
  p_same <- 1 / 4 + 3 / 4 * exp(-4 * d / 3)
  p_other <- 1 / 4 - 1 / 4 * exp(-4 * d / 3)

  P <- matrix(
    p_other,
    nrow = 4,
    ncol = 4,
    dimnames = list(from = nucleotides, to = nucleotides)
  )

  diag(P) <- p_same
  P
}

simulate_jc_sequence <- function(ancestor, d) {
  P <- jc_P(d)

  map_chr(
    ancestor,
    function(base) {
      sample(nucleotides, size = 1, prob = P[base, ])
    }
  )
}
```

## Today

By the end, you should be able to:

1. Explain what makes a sequence a **Markov chain**.
2. Represent a finite-state Markov chain with a transition matrix.
3. Simulate Markov chains in R.
4. Interpret multi-step and long-run behavior.
5. Recognize Brownian diffusion as a continuous-state Markov model.
6. Simulate geographic diffusion with `mvtnorm::rmvnorm()`.

## From the previous lecture

Last time, we simulated categorical values using something like

```{r}
sample(
  c("Dry", "Wet"),
  size = 10,
  replace = TRUE,
  prob = c(.7, .3)
)
```

Every draw uses the **same probabilities**.

. . .

But many processes have memory:

> If today is wet, does that change the probability that tomorrow is wet?

## Independent versus dependent simulation

Suppose weather is either **Dry** or **Wet**.

An independent model might say

$$
Prob(\text{wet tomorrow}) = 0.3
$$
no matter what happened today.

A dependent model might say

$$ 
Prob(\text{wet tomorrow given wet today}) \neq Prob(\text{wet tomorrow given dry today})
$$

The probability used for the next state depends on the current state.

## The Markov idea

A **Markov chain** is a sequence of states

$$
X_0, X_1, X_2, \ldots
$$

where the next state depends on the current state.

Examples:

- today's weather $\rightarrow$ tomorrow's weather
- a customer's current status $\rightarrow$ next month's status
- an animal's current habitat $\rightarrow$ next habitat
- a nucleotide $\rightarrow$ a nucleotide later in evolutionary time

## The Markov property

The defining assumption is

$$
P(X_{t+1} \mid X_t, X_{t-1}, \ldots, X_0)
=
P(X_{t+1} \mid X_t).
$$

Once we know the **present state**, the earlier history gives no additional
information about the next state.

::: {.callout-warning}
Markov does **not** mean independent.

Usually $X_{t+1}$ and $X_t$ are strongly dependent.
:::

## A two-state weather model

Suppose

- if today is **Dry**, tomorrow is Dry with probability $0.85$
- if today is **Dry**, tomorrow is Wet with probability $0.15$
- if today is **Wet**, tomorrow is Dry with probability $0.40$
- if today is **Wet**, tomorrow is Wet with probability $0.60$

We can store all four probabilities together.

## Transition matrix

::: panel-tabset

## Math

$$
P =
\begin{pmatrix}
0.85 & 0.15 \\
0.40 & 0.60
\end{pmatrix}
$$

## In R

```{r}
states <- c("Dry", "Wet")

P_weather <- matrix(
  c(.85, .15,
    .40, .60),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(from = states, to = states)
)
```
:::



Rows describe the state **now** and columns describe the state **next**.

| From / To | Dry | Wet |
|---|---:|---:|
| Dry | 0.85 | 0.15 |
| Wet | 0.40 | 0.60 |

::: {.callout-note}
Every row of a transition matrix sums to 1.
:::

## Store the transition matrix in R

```{r}
P_weather
```

Because we gave the rows and columns names, we can ask for the probabilities
associated with the current state.

```{r}
P_weather["Dry", ]
P_weather["Wet", ]
```

## Simulate one transition

Suppose today is Dry.

```{r}
current <- "Dry"

sample(
  states,
  size = 1,
  prob = P_weather[current, ]
)
```

The key difference from our previous simulation lecture:

> `prob =` now depends on the current state.

## Simulate an entire Markov chain

```{r}
#| eval: false
simulate_chain <- function(P, initial, n_steps = 100) {
  states <- rownames(P)

  chain <- character(n_steps + 1)
  chain[1] <- initial

  for (t in seq_len(n_steps)) {
    chain[t + 1] <- sample(
      states,
      size = 1,
      prob = P[chain[t], ]
    )
  }

  chain
}
```

At every step:

1. look at the current state,
2. retrieve the corresponding row of $P$,
3. simulate the next state.

## One simulated weather sequence

::: panel-tabset
## plot

```{r}
#| echo: false
weather_path <- simulate_chain(
  P_weather,
  initial = "Dry",
  n_steps = 80
)

tibble(
  day = 0:80,
  state = factor(weather_path, levels = states)
) |>
  ggplot(aes(x = day, y = state, group = 1)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Day",
    y = NULL,
    title = "One realization of the Markov chain"
  )
```

## code

```{r}
#| eval: false
weather_path <- simulate_chain(
  P_weather,
  initial = "Dry",
  n_steps = 80
)

tibble(
  day = 0:80,
  state = factor(weather_path, levels = states)
) |>
  ggplot(aes(x = day, y = state, group = 1)) +
  geom_line() +
  geom_point()
```
:::

## Exercise 1

Modify the transition matrix so weather is **more persistent**:

- Dry $\rightarrow$ Dry has probability $0.95$
- Wet $\rightarrow$ Wet has probability $0.90$

Then:

1. construct the new transition matrix,
2. simulate 80 days,
3. compare the sequence with the previous simulation.

What changes visually?

## More than one step ahead

The transition matrix tells us what happens **one step** ahead.

What is the probability of going from Dry today to Wet **two days from now**?

There are two possible paths:

$$
\text{Dry}\rightarrow\text{Dry}\rightarrow\text{Wet}
$$

or

$$
\text{Dry}\rightarrow\text{Wet}\rightarrow\text{Wet}.
$$

So

$$
P(\text{Dry}\rightarrow\text{Wet in 2 steps})
=
(.85)(.15)+(.15)(.60)
=
0.2175.
$$

## Matrix multiplication does the bookkeeping

```{r}
P_weather %*% P_weather
```

The resulting matrix contains all **two-step transition probabilities**.

More generally,

$$
P^k
$$

contains the $k$-step transition probabilities.

::: {.callout-warning}
In R, `P^2` squares each entry separately.

Use `%*%` for matrix multiplication.
:::

## Computing $P^k$

::: panel-tabset 

## Manually 
```{r}
#| eval: false
matrix_power <- function(P, k) {
  out <- diag(nrow(P))

  for (i in seq_len(k)) {
    out <- out %*% P
  }

  dimnames(out) <- dimnames(P)
  out
}
```

## expm

```{r}
#| message: FALSE
#| warning: false
library(expm)

P_weather %^% 10
```
:::

For example:

```{r}
matrix_power(P_weather, 10)
```

Notice how the two rows are becoming similar.

## Long-run behavior

Simulate the weather chain for a long time.

```{r}
long_chain <- simulate_chain(
  P_weather,
  initial = "Dry",
  n_steps = 10000
)

prop.table(table(long_chain))
```

Even though the weather is dependent from day to day, its **long-run fraction**
of time in each state can settle down.

## Stationary distribution

A distribution $\pi$ is **stationary** when

$$
\pi = \pi P.
$$

Interpretation:

> If the chain currently follows distribution $\pi$, one more transition
> leaves the distribution unchanged.

We can find it computationally by repeatedly applying the transition matrix.

```{r}
pi <- c(.5, .5)

for (i in 1:100) {
  pi <- as.numeric(pi %*% P_weather)
}

set_names(pi, states)
```

## Simulation and the stationary distribution

For this chain, the long-run proportions from simulation should be close to
the stationary probabilities.

```{r}
sim_prop <- prop.table(table(long_chain))

tibble(
  state = states,
  stationary = pi,
  simulated = as.numeric(sim_prop)
)
```

::: {.callout-note}
This connection between a Markov chain and its stationary distribution is the
central idea behind **Markov chain Monte Carlo (MCMC)**.
:::

## Definitions

::: callout-note
## Definition
**Irreducible**: Every state can be reached from every other state with positive probability in some finite number of steps.
:::

::: callout-note
## Definition
**Recurrent**: Starting from any state, the chain returns to that state eventually with probability 1. 
:::

::: callout-note
## Definition
**Positive recurrent**: The chain is recurrent and the expected amount of time until returning to every state is finite.
:::

## When are we guaranteed stationarity?

For a countable-state Markov chain,

$$
\text{irreducible + positive recurrent}
\implies
\text{unique stationary distribution}.
$$

The contrapositive is also true:

$$
\text{no unique stationary distribution}
\implies
\text{not irreducible or not positive recurrent}.
$$

But the **converse is not true**:

$$
\text{unique stationary distribution}
\nRightarrow  
\text{irreducible + positive recurrent}.
$$

A related example is the simple symmetric random walk on the integers:

$$
X_{t+1} =
\begin{cases}
X_t + 1, & \text{with probability } 1/2,\\
X_t - 1, & \text{with probability } 1/2.
\end{cases}
$$

This chain is:

irreducible and recurrent, but **not positive recurrent**.

In fact, the expected return time to any state is infinite. Therefore it has **no stationary probability distribution**. So recurrence alone is not enough: we need **positive** recurrence.

# Continuous states

## Markov models need not have discrete states

So far the state was one of two labels.

But the state could instead be a pair of continuous coordinates:

$$
X_t =
\begin{pmatrix}
x_t \\
y_t
\end{pmatrix}.
$$

For example, $X_t$ could be the geographic position of an organism at time
$t$.

## Brownian diffusion

A simple diffusion model says

Position at time $t + \Delta t$ given position at time $t$ is drawn from a multivariate normal distribution (`mvtnorm::rmvnorm(x_t, mean = c(0, 0), Sigma)`), centered on the position at time $t$. 


<!-- $$ -->
<!-- X_{t+\Delta t}\mid X_t -->
<!-- \sim -->
<!-- N_2(X_t,\Sigma \Delta t). -->
<!-- $$ -->

Equivalently, we can write 

$$
X_{t+\Delta t} = X_t + \epsilon_t,
\qquad
\epsilon_t \sim N_2(0,\Sigma\Delta t).
$$

The next location depends on the current location, but not on the entire path.

That is the Markov property again.

## Simulating a 2D step with `rmvnorm()`

The `mvtnorm` package generates multivariate normal random variables.

```{r}
Sigma_geo
```

```{r}
rmvnorm(
  n = 5,
  mean = c(0, 0),
  sigma = Sigma_geo
)
```

Each row is a possible 2D displacement.

The covariance matrix controls the direction and magnitude of spread.

## Simulating geographic diffusion

We will measure location relative to the starting point:

- $x$: kilometers east/west
- $y$: kilometers north/south

```{r}
#| eval: false
simulate_brownian <- function(n_steps = 100,
                              start = c(0, 0),
                              sigma = diag(2),
                              dt = 1) {

  locations <- matrix(
    NA,
    nrow = n_steps + 1,
    ncol = 2
  )

  locations[1, ] <- start

  for (t in 1:n_steps) {

    move <- rmvnorm(
      1,
      mean = c(0, 0),
      sigma = sigma * dt
    )

    locations[t + 1, ] <- locations[t, ] + move
  }

  tibble(
    step = 0:n_steps,
    east_km = locations[, 1],
    north_km = locations[, 2]
  )
}
```

## One diffusion path

::: panel-tabset
## plot

```{r}
#| echo: false
brownian_path <- simulate_brownian(
  n_steps = 100,
  sigma = Sigma_geo
)

brownian_path |>
  ggplot(aes(x = east_km, y = north_km)) +
  geom_path() +
  geom_point(data = slice_head(brownian_path, n = 1), size = 3) +
  geom_point(data = slice_tail(brownian_path, n = 1), size = 3) +
  coord_equal() +
  labs(
    x = "Kilometers east/west",
    y = "Kilometers north/south",
    title = "One Brownian diffusion path"
  )
```

## code

```{r}
#| eval: false
brownian_path <- simulate_brownian(
  n_steps = 100,
  sigma = Sigma_geo
)

brownian_path |>
  ggplot(aes(x = east_km, y = north_km)) +
  geom_path() +
  coord_equal()
```
:::

## Many possible geographic histories

::: panel-tabset
## plot

```{r}
#| echo: false
brownian_paths <- map_dfr(
  1:25,
  function(id) {
    simulate_brownian(
      n_steps = 100,
      sigma = Sigma_geo
    ) |>
      mutate(path = factor(id))
  }
)

brownian_paths |>
  ggplot(aes(
    x = east_km,
    y = north_km,
    group = path
  )) +
  geom_path(alpha = .35) +
  coord_equal() +
  labs(
    x = "Kilometers east/west",
    y = "Kilometers north/south",
    title = "25 simulated diffusion histories"
  )
```

## code

```{r}
#| eval: false
brownian_paths <- map_dfr(
  1:25,
  function(id) {
    simulate_brownian(
      n_steps = 100,
      sigma = Sigma_geo
    ) |>
      mutate(path = factor(id))
  }
)

brownian_paths |>
  ggplot(aes(east_km, north_km, group = path)) +
  geom_path(alpha = .35) +
  coord_equal()
```
:::

## What does $\Sigma$ do?

For

$$
\Sigma =
\begin{pmatrix}
1.2 & 0.45 \\
0.45 & 0.7
\end{pmatrix},
$$

- $1.2$: east/west variance per unit time
- $0.7$: north/south variance per unit time
- $0.45$: correlation between east/west and north/south displacement

::: {.callout-note}
For real geographic analyses over nontrivial distances, work in an appropriate
projected coordinate system rather than directly adding normal noise to
longitude and latitude.
:::

## Exercise 2

Change the diffusion model so spread is much greater east/west than
north/south.

1. Choose a new covariance matrix $\Sigma$.
2. Simulate 25 paths.
3. Plot them.
4. Explain how your choice of $\Sigma$ appears in the plot.

Optional: set the off-diagonal covariance to zero and compare.

# Bonus / next time

## Discrete time versus continuous time

Our weather chain changed at fixed steps:

$$
t=0,1,2,3,\ldots
$$

But many processes evolve continuously.

For nucleotide evolution, a substitution can occur at **any point in time**.

This leads to a **continuous-time Markov chain**.

## A biological example

DNA is written using four nucleotide states:

$$
\{A,C,G,T\}.
$$

Suppose an ancestral sequence contains

```text
A C G T T A C G A
```

After evolutionary time passes, a descendant might contain

```text
A C A T T G C G A
```

We want a model for

$$
P(\text{descendant nucleotide}\mid
\text{ancestral nucleotide},\text{time}).
$$

<!-- ## Jukes--Cantor 1969 -->

<!-- The Jukes--Cantor model is the simplest standard model of nucleotide -->
<!-- substitution. -->

<!-- Its main assumptions are: -->

<!-- 1. the four states are $A,C,G,T$, -->
<!-- 2. all three possible substitutions away from a nucleotide occur at the same rate, -->
<!-- 3. the long-run nucleotide frequencies are all $1/4$, -->
<!-- 4. different sites are modeled independently in the basic model. -->

<!-- This is a continuous-time Markov chain. -->

<!-- ## From transition probabilities to transition rates -->

<!-- For continuous-time chains we use a **rate matrix** $Q$. -->

<!-- We will parameterize the model so $\mu$ is the total substitution rate away -->
<!-- from the current nucleotide: -->

<!-- $$ -->
<!-- Q -->
<!-- = -->
<!-- \mu -->
<!-- \begin{pmatrix} -->
<!-- -1 & 1/3 & 1/3 & 1/3\\ -->
<!-- 1/3 & -1 & 1/3 & 1/3\\ -->
<!-- 1/3 & 1/3 & -1 & 1/3\\ -->
<!-- 1/3 & 1/3 & 1/3 & -1 -->
<!-- \end{pmatrix}. -->
<!-- $$ -->

<!-- ::: {.callout-warning} -->
<!-- A rate matrix is **not** a probability matrix. -->

<!-- Its rows sum to 0, not 1. -->
<!-- ::: -->

<!-- ## From rates to probabilities -->

<!-- For a continuous-time Markov chain, -->

<!-- $$ -->
<!-- P(t)=e^{Qt}. -->
<!-- $$ -->

<!-- You do **not** need to compute the matrix exponential by hand. -->

<!-- For Jukes--Cantor we get a simple closed form. -->

<!-- Let -->

<!-- $$ -->
<!-- d=\mu t -->
<!-- $$ -->

<!-- be the expected number of substitutions per site along the branch. -->

<!-- ## Jukes--Cantor transition probabilities -->

<!-- Define -->

<!-- $$ -->
<!-- s(d)=\frac14+\frac34e^{-4d/3} -->
<!-- $$ -->

<!-- and -->

<!-- $$ -->
<!-- r(d)=\frac14-\frac14e^{-4d/3}. -->
<!-- $$ -->

<!-- Then -->

<!-- $$ -->
<!-- P(d)= -->
<!-- \begin{pmatrix} -->
<!-- s&r&r&r\\ -->
<!-- r&s&r&r\\ -->
<!-- r&r&s&r\\ -->
<!-- r&r&r&s -->
<!-- \end{pmatrix}. -->
<!-- $$ -->

<!-- - $s(d)$: probability the nucleotide is the same after time $d$ -->
<!-- - $r(d)$: probability of changing to any **specific** other nucleotide -->

<!-- ## Check the limiting cases -->

<!-- When $d=0$, -->

<!-- $$ -->
<!-- s(0)=1,\qquad r(0)=0. -->
<!-- $$ -->

<!-- So nothing has had time to change. -->

<!-- . . . -->

<!-- As $d\rightarrow\infty$, -->

<!-- $$ -->
<!-- s(d)\rightarrow\frac14, -->
<!-- \qquad -->
<!-- r(d)\rightarrow\frac14. -->
<!-- $$ -->

<!-- Eventually the process forgets where it started. -->

<!-- The stationary distribution is -->

<!-- $$ -->
<!-- \pi=(1/4,1/4,1/4,1/4). -->
<!-- $$ -->

<!-- ## Implement $P(d)$ in R -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- jc_P <- function(d) { -->
<!--   p_same <- 1 / 4 + 3 / 4 * exp(-4 * d / 3) -->
<!--   p_other <- 1 / 4 - 1 / 4 * exp(-4 * d / 3) -->

<!--   P <- matrix( -->
<!--     p_other, -->
<!--     nrow = 4, -->
<!--     ncol = 4, -->
<!--     dimnames = list( -->
<!--       from = nucleotides, -->
<!--       to = nucleotides -->
<!--     ) -->
<!--   ) -->

<!--   diag(P) <- p_same -->
<!--   P -->
<!-- } -->
<!-- ``` -->

<!-- For example: -->

<!-- ```{r} -->
<!-- round(jc_P(.5), 3) -->
<!-- ``` -->

<!-- ## How substitution probabilities change with time -->

<!-- ::: panel-tabset -->
<!-- ## plot -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- jc_curve <- tibble( -->
<!--   d = seq(0, 4, length.out = 300), -->
<!--   same = 1 / 4 + 3 / 4 * exp(-4 * d / 3), -->
<!--   different = 3 / 4 * (1 - exp(-4 * d / 3)) -->
<!-- ) |> -->
<!--   pivot_longer( -->
<!--     cols = c(same, different), -->
<!--     names_to = "outcome", -->
<!--     values_to = "probability" -->
<!--   ) -->

<!-- jc_curve |> -->
<!--   ggplot(aes(x = d, y = probability, linetype = outcome)) + -->
<!--   geom_line(linewidth = 1.1) + -->
<!--   labs( -->
<!--     x = "Expected substitutions per site, d", -->
<!--     y = "Probability", -->
<!--     linetype = NULL, -->
<!--     title = "Jukes–Cantor: observed differences saturate" -->
<!--   ) -->
<!-- ``` -->

<!-- ## code -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- tibble( -->
<!--   d = seq(0, 4, length.out = 300), -->
<!--   same = 1 / 4 + 3 / 4 * exp(-4 * d / 3), -->
<!--   different = 3 / 4 * (1 - exp(-4 * d / 3)) -->
<!-- ) |> -->
<!--   pivot_longer( -->
<!--     cols = c(same, different), -->
<!--     names_to = "outcome", -->
<!--     values_to = "probability" -->
<!--   ) |> -->
<!--   ggplot(aes(d, probability, linetype = outcome)) + -->
<!--   geom_line() -->
<!-- ``` -->
<!-- ::: -->

<!-- ## Simulate one nucleotide along a branch -->

<!-- Suppose the ancestral nucleotide is `A` and the branch length is $d=0.5$. -->

<!-- ```{r} -->
<!-- P <- jc_P(.5) -->

<!-- P["A", ] -->
<!-- ``` -->

<!-- Then simulate the descendant nucleotide: -->

<!-- ```{r} -->
<!-- sample( -->
<!--   nucleotides, -->
<!--   size = 1, -->
<!--   prob = P["A", ] -->
<!-- ) -->
<!-- ``` -->

<!-- This is exactly the same computational pattern as the weather Markov chain. -->

<!-- ## Simulate an entire DNA sequence -->

<!-- ```{r} -->
<!-- ancestor <- sample( -->
<!--   nucleotides, -->
<!--   size = 1000, -->
<!--   replace = TRUE -->
<!-- ) -->

<!-- descendant <- simulate_jc_sequence( -->
<!--   ancestor, -->
<!--   d = .5 -->
<!-- ) -->

<!-- mean(ancestor != descendant) -->
<!-- ``` -->

<!-- The theoretical probability that a site differs is -->

<!-- $$ -->
<!-- p_{\text{different}} -->
<!-- = -->
<!-- \frac34\left(1-e^{-4d/3}\right). -->
<!-- $$ -->

<!-- For $d=0.5$: -->

<!-- ```{r} -->
<!-- 3 / 4 * (1 - exp(-4 * .5 / 3)) -->
<!-- ``` -->

<!-- ## Why observed differences underestimate change -->

<!-- A site can change more than once: -->

<!-- $$ -->
<!-- A\rightarrow G\rightarrow A -->
<!-- $$ -->

<!-- or -->

<!-- $$ -->
<!-- A\rightarrow C\rightarrow T. -->
<!-- $$ -->

<!-- If we compare only the ancestor and descendant: -->

<!-- - the first site looks unchanged, -->
<!-- - the second site shows only one difference, -->

<!-- even though both experienced multiple substitutions. -->

<!-- This is why the raw fraction of different sites is not the same thing as the -->
<!-- number of substitutions per site. -->

<!-- ## The Jukes--Cantor correction -->

<!-- Let $p$ be the observed fraction of sites that differ. -->

<!-- Under Jukes--Cantor, -->

<!-- $$ -->
<!-- p -->
<!-- = -->
<!-- \frac34\left(1-e^{-4d/3}\right). -->
<!-- $$ -->

<!-- Solving for $d$, -->

<!-- $$ -->
<!-- \boxed{ -->
<!-- d -->
<!-- = -->
<!-- -\frac34 -->
<!-- \log\left(1-\frac43p\right) -->
<!-- } -->
<!-- $$ -->

<!-- This corrects for unobserved multiple substitutions. -->

<!-- ## Example -->

<!-- Suppose two aligned sequences differ at 30% of sites. -->

<!-- The naive distance is -->

<!-- $$ -->
<!-- p=0.30. -->
<!-- $$ -->

<!-- The Jukes--Cantor corrected distance is -->

<!-- ```{r} -->
<!-- p <- .30 -->
<!-- -3 / 4 * log(1 - 4 * p / 3) -->
<!-- ``` -->

<!-- So the model estimates about $0.383$ substitutions per site rather than -->
<!-- $0.30$. -->

<!-- ## Exercise 3 -->

<!-- Simulate nucleotide evolution under Jukes--Cantor. -->

<!-- 1. Generate an ancestral sequence of 2000 nucleotides. -->
<!-- 2. Simulate descendants using $d=0.1$, $d=0.5$, and $d=2$. -->
<!-- 3. For each value of $d$, calculate the fraction of sites that differ. -->
<!-- 4. Compare each result with -->

<!-- $$ -->
<!-- \frac34\left(1-e^{-4d/3}\right). -->
<!-- $$ -->

<!-- What happens to the observed fraction of differences as $d$ becomes very -->
<!-- large? -->

<!-- ## One idea, several models -->

<!-- | Model | State | Next state depends on | Transition rule | -->
<!-- |---|---|---|---| -->
<!-- | Weather chain | Dry / Wet | current weather | row of $P$ | -->
<!-- | Brownian diffusion | $(x,y)$ | current location | multivariate normal | -->
<!-- | Jukes--Cantor | A / C / G / T | current nucleotide + elapsed time | $P(d)$ | -->

<!-- The common idea is the **Markov property**: -->

<!-- > Given the present, the next transition does not require the full past. -->

<!-- ## Takeaways -->

<!-- 1. Markov chains let us simulate **dependent sequences**. -->
<!-- 2. A transition matrix summarizes one-step probabilities for finite-state chains. -->
<!-- 3. Repeated transitions reveal multi-step and stationary behavior. -->
<!-- 4. Brownian diffusion extends the Markov idea to continuous geographic states. -->
<!-- 5. Jukes--Cantor extends it to continuous time for nucleotide substitution. -->
<!-- 6. Simulation turns all of these models into explicit algorithms. -->

<!-- ## Sources -->

<!-- - Jukes, T. H. & Cantor, C. R. (1969). *Evolution of Protein Molecules*. In H. N. Munro (Ed.), *Mammalian Protein Metabolism*, Vol. III, pp. 21--132. Academic Press. DOI: 10.1016/B978-1-4832-3211-9.50009-7 -->
<!-- - `mvtnorm` package documentation: <https://CRAN.R-project.org/package=mvtnorm> -->
<!-- - Background on nucleotide Markov models: <https://cran.r-project.org/web/packages/phangorn/vignettes/AdvancedFeatures.html> -->
