---
title: "Spatial data"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---


```{r setup}
#| include: false
# terra is attached FIRST so that dplyr/tidyr versions of shared names
# (e.g. extract, select) win. We call terra::extract() explicitly below.
library(terra)
library(tidyverse)
library(sf)
library(patchwork)

set.seed(323)
theme_set(theme_minimal(base_size = 18))
```

## Today

By the end, you should be able to:

1. Tell apart the two ways of storing spatial data: **vector** (objects) and **raster** (grids).
2. Wrangle and map **simple features** with `sf`, `dplyr`, and `geom_sf()`.
3. Load, transform, and summarize a **raster** with `terra`.
4. Explain why spatial data need extra care: **coordinate systems, scale, and neighbors**.
5. Use **simulation** to see what spatial dependence looks like, and to test for it.


## Spatial data = ordinary data + geometry

Most data frames have rows and columns.

Spatial data frames also record **where each row lives**.

| ordinary variable | spatial variable |
|---|---|
| population | county boundary |
| restaurant rating | restaurant location |
| traffic volume | road segment |
| temperature | weather station location |

::: {.callout-note}
A spatial object is usually ordinary data plus a special **geometry** component.
:::

## Two ways to represent space

::: {.columns}
::: {.column width="50%"}
**Vector data = objects**

- **point**: one location
- **line**: a path or route
- **polygon**: an area or boundary

One row = one object.

Examples: stores, roads, counties.

In R: `sf`.
:::

::: {.column width="50%"}
**Raster data = grid**

- space is divided into equal-sized cells
- each cell stores a value

One value per cell, like pixels in a photo.

Examples: elevation, temperature, satellite imagery.

In R: `terra` (or `stars`).
:::
:::

::: {.callout-note}
Rule of thumb: **discrete things** → vector. **Quantities that vary continuously over space** → raster.
:::

## Exercise 1: Vector or raster?

For each example: is it vector (point, line, or polygon) or raster? What is **one row** (or one cell)?

1. Locations of every coffee shop in Durham.
2. County-level unemployment rates in North Carolina.
3. Temperature measured at 75 weather stations.
4. GPS locations from a tagged sea turtle, every hour.
5. A satellite image measuring vegetation intensity.

## Common spatial file formats

| Format | Common use | R package |
|---|---|---|
| Shapefile: `.shp` + friends | older vector GIS data | `sf` |
| GeoPackage: `.gpkg` | modern vector GIS data | `sf` |
| GeoJSON: `.geojson` | web maps, APIs | `sf` |
| CSV with lon/lat columns | point locations | `readr`, then `sf` |
| GeoTIFF: `.tif` | raster/grid data | `terra` |

::: {.callout-tip}
For vector data, ask: **can I read this with `sf::st_read()`?**
For raster data, ask: **can I read this with `terra::rast()`?**
:::

# Part 1: Vector data with `sf`

## Simple features: the `sf` package

`sf` stands for **simple features**. It gives us spatial data frames that work with `dplyr`, `ggplot2`, and pipes.

```{r}
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

nc |> select(NAME, BIR74, SID74)
```

We only selected three columns... but a fourth one, `geometry`, came along anyway.

## What is inside an `sf` object?

Three ingredients:

1. **ordinary columns** such as `NAME`, `BIR74`, `SID74`
2. one special **`geometry` column** storing the shapes (it is "sticky": `select()` keeps it)
3. metadata describing the **coordinate reference system**, or CRS

```{r}
class(nc)
st_geometry_type(nc)[1:3]
```

You generally leave `geometry` alone and work with it through `st_*()` functions: `st_area()`, `st_buffer()`, `st_join()`, ...

## First map: `geom_sf()`

```{r}
ggplot(nc) +
  geom_sf() +
  labs(title = "North Carolina counties")
```

`geom_sf()` knows how to draw the geometry column, so we do **not** write `aes(x = ..., y = ...)`.

## Mapping an attribute

```{r}
nc |>
  mutate(sids_rate = 1000 * SID74 / BIR74) |>
  ggplot() +
  geom_sf(aes(fill = sids_rate), color = "white") +
  scale_fill_viridis_c() +
  labs(
    title = "Sudden infant death syndrome (SIDS), 1974–78",
    fill = "deaths per\n1,000 births"
  )
```

For a polygon map, `aes(fill = variable)` creates a **choropleth**.

::: {.callout-tip}
Map **rates**, not raw counts: bigger counties produce bigger counts. And beware: counties with few births have **noisy** rates, so the most extreme colors are often the smallest counties.
:::

## Thought exercise: Build a point map

Suppose we have:

```{r}
#| eval: false
coffee <- tibble(
  name = c("A", "B", "C"),
  longitude = c(-78.91993, -78.93016, -78.93805),
  latitude = c(36.00932, 36.01158, 36.00346),
  rating = c(4.6, 4.8, 5.0)
)
```

What two steps are needed to turn this into a spatial map?

::: {.fragment}
1. Convert the longitude/latitude columns into a geometry column.
2. Plot the resulting `sf` object with `geom_sf()`.

But to do step 1, R needs one more piece of information... (next slide)
:::

## Coordinate reference systems

A **CRS** tells R what the coordinates mean.

```{r}
st_crs(nc)$input
st_is_longlat(nc)
```

Two broad cases:

- **longitude / latitude**: angles on a round Earth, in degrees
- **projected coordinates**: the Earth flattened onto a plane, measured in meters or feet

::: {.callout-warning}
`sf` always reads coordinates as **(x, y) = (longitude, latitude)**. The pair `(36, -79)` read that way is in Antarctica, not Durham!
:::

## From longitude/latitude columns to `sf`

```{r}
places <- tibble(
  place = c("Duke", "UNC", "NC State"),
  lon = c(-78.9382, -79.0469, -78.6821),
  lat = c(36.0014, 35.9049, 35.7847)
)

places_sf <- places |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

`crs = 4326` is the code for plain GPS longitude/latitude (WGS 84), which is what most web and phone data use. Once converted, spatial layers stack like ordinary `ggplot2` layers:

```{r}
#| fig-height: 3
ggplot(nc) +
  geom_sf(fill = "gray95", color = "white") +
  geom_sf(data = places_sf, size = 4)
```

## The most common CRS mistake

::: {.callout-warning}
Do not treat longitude/latitude degrees as ordinary distances. One degree of longitude is about 56 miles in Durham, but 0 miles at the North Pole.
:::

For distances, areas, and buffers, transform to a suitable **projected** CRS first:

```{r}
nc_projected <- nc |>
  st_transform(2264) # NAD83 / North Carolina, units = US feet
```

| Function | Meaning |
|---|---|
| `st_set_crs()` | **label** coordinates that are already in a CRS (numbers unchanged) |
| `st_transform()` | **convert** coordinates to another CRS (numbers change) |

<!-- ## Exercise 3: What is wrong with this buffer? -->

<!-- Why might this code be wrong? -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- coffee_sf |> -->
<!--   st_buffer(dist = 1) -->
<!-- ``` -->

<!-- What does `1` mean? -->

<!-- ::: {.fragment} -->
<!-- It depends on the CRS. If the data are longitude/latitude, `1` is not "one mile" or "one kilometer." -->
<!-- ::: -->

<!-- ::: {.fragment} -->
<!-- **Better workflow:** `st_transform()` to a projected CRS, then buffer using that CRS's units (for EPSG 2264, feet). -->
<!-- ::: -->

## Wrangling spatial data with `dplyr`

Most `dplyr` verbs work on `sf` objects, and the geometry comes along for the ride.

```{r}
nc_west <- nc |>
  filter(NAME %in% c("Buncombe", "Watauga", "Ashe", "Avery", "Mitchell")) |>
  mutate(region = "mountains") |>
  select(NAME, region, BIR74, SID74)
```

When geometry is irrelevant, drop it explicitly (it makes tables smaller and code faster):

```{r}
nc |>
  st_drop_geometry() |>
  select(NAME, BIR74, SID74) |>
  slice(1:5)
```

Spatial data are still **data**; use ordinary data tools when appropriate.

## Deriving spatial variables

Geometry can generate new variables: area, length, distance, centroid, ...

```{r}
nc_features <- nc_projected |>
  mutate(area_sq_miles = as.numeric(st_area(geometry)) / 5280^2)

nc_features |>
  st_drop_geometry() |>
  select(NAME, area_sq_miles) |>
  arrange(desc(area_sq_miles)) |>
  slice(1:5)
```

`st_area()` returns square feet here because EPSG 2264 is in feet; dividing by 5280² converts to square miles.

<!-- ::: {.callout-note} -->
<!-- `st_centroid()` gives a "center point" of each polygon. Handy, but for oddly shaped polygons it can fall **outside** the polygon. -->
<!-- ::: -->

## Spatial predicates

A **spatial predicate** asks a TRUE/FALSE geometry question.

| Predicate | Question |
|---|---|
| `st_intersects()` | Do these geometries overlap or touch? |
| `st_within()` | Is A inside B? |
| `st_contains()` | Does A contain B? |
| `st_touches()` | Do their boundaries touch (without overlapping)? |
| `st_is_within_distance()` | Are they close enough? |

These functions turn geometry into **relationships between rows**.

```{r}
#| echo: false
#| eval: false
library(tidyverse)
library(sf)

coffee <- tibble(
  name = c("A", "B", "C"),
  longitude = c(-78.91993, -78.93016, -78.93805),
  latitude = c(36.00932, 36.01158, 36.00346),
  rating = c(4.4, 4.5, 5.0)
)

# 1. Make both layers sf objects, then project to NC State Plane (units = feet)
coffee_sf <- coffee |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(2264)

chapel <- tibble(name = "Duke Chapel", lon = -78.9401, lat = 36.0019) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(2264)

# 2. Ask the yes/no question
st_is_within_distance(coffee_sf, chapel, dist = 5280, sparse = FALSE)
#>       [,1]
#> [1,] FALSE   # A
#> [2,]  TRUE   # B
#> [3,]  TRUE   # C

# 3. Use the answer like any logical vector
coffee_sf |>
  mutate(near_chapel = st_is_within_distance(geometry, chapel,
                                             dist = 5280, sparse = FALSE)[, 1])
```


## Spatial joins

An ordinary join matches rows using a key column. A **spatial join** matches rows using a spatial predicate.

```{r}
places_for_join <- places_sf |>
  st_transform(st_crs(nc)) # both layers must share a CRS

places_with_county <- st_join(
  places_for_join,
  nc |> select(NAME),
  join = st_within
)

places_with_county
```

Question answered: **which county contains each point?**

## Exercise 2

Which coffee shop is within 1 mile of Duke Chapel?

```{r}
chapel <- tibble(name = "Duke Chapel", lon = -78.9401, lat = 36.0019) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(2264)
```


<!-- You have coffee shop points and neighborhood polygons. What operation answers each question? -->

<!-- 1. Which neighborhood contains each coffee shop? -->
<!-- 2. Which shops are within 1 mile of campus? -->
<!-- 3. Which neighborhoods touch one another? -->
<!-- 4. How many shops are in each neighborhood? -->

<!-- ::: {.notes} -->
<!-- Answers: st_join(shops, neighborhoods, join = st_within); st_is_within_distance() or a buffer in a projected CRS; st_touches(); spatial join, then count() by neighborhood. -->
<!-- ::: -->

## Buffers: all locations within a distance

```{r}
#| output-location: column
places_ncft <- places_sf |>
  st_transform(2264)

buffers <- places_ncft |>
  st_buffer(dist = 30 * 5280) # 30 miles, in feet

ggplot() +
  geom_sf(data = nc_projected,
          fill = "gray95", color = "white") +
  geom_sf(data = buffers,
          fill = NA, linewidth = 1) +
  geom_sf(data = places_ncft, size = 3) +
  labs(title = "30-mile buffers")
```

## Spatial summaries and dissolving boundaries

Which counties intersect any campus buffer? Then, `group_by()` + `summarize()` **merges the shapes** within each group.

::: panel-tabset

## Plots

```{r}
#| echo: false
nc_near <- nc_projected |>
  mutate(near_campus = lengths(st_intersects(geometry, st_union(buffers))) > 0)

nc_dissolved <- nc_near |>
  group_by(near_campus) |>
  summarize(n_counties = n())

p1 <- ggplot(nc_near) + geom_sf(aes(fill = near_campus), color = "white") +
  labs(title = "100 counties")
p2 <- ggplot(nc_dissolved) + geom_sf(aes(fill = near_campus), color = "white") +
  labs(title = "2 dissolved regions")
p1 + p2 + plot_layout(guides = "collect")
```

## Code
```{r}
#| eval: false
nc_near <- nc_projected |>
  mutate(near_campus = lengths(st_intersects(geometry, st_union(buffers))) > 0)

nc_dissolved <- nc_near |>
  group_by(near_campus) |>
  summarize(n_counties = n())

p1 <- ggplot(nc_near) + geom_sf(aes(fill = near_campus), color = "white") +
  labs(title = "100 counties")
p2 <- ggplot(nc_dissolved) + geom_sf(aes(fill = near_campus), color = "white") +
  labs(title = "2 dissolved regions")
p1 + p2 + plot_layout(guides = "collect")
```
::: 

## A spatial debugging checklist

Spatial bugs are usually **data representation bugs**, not syntax bugs. When something fails or looks strange:

1. Are both objects `sf` objects?
2. Do they have the **same CRS**?
3. Am I measuring distance or area in a **projected** CRS?
4. Are the geometries valid?
5. Did a spatial join create **one-to-many** matches?
6. Am I mapping a **count** when I need a **rate**?

```{r}
#| eval: false
st_crs(x)            # what coordinate system?
st_geometry_type(x)  # points, lines, polygons?
st_is_valid(x)       # any broken shapes?
st_bbox(x)           # does the extent look like the right place?
ggplot(x) + geom_sf()  # always look at it!
```

A beautiful map can still encode a poor analysis.

# Part 2: Raster data with `terra`

## A raster is a grid of values

Here is a toy raster: every cell gets one value. Most raster operations transform **every cell at once**.


::: panel-tabset

## Plot
```{r}
#| echo: false
raster_grid <- crossing(
  x = seq(0, 1, length.out = 35),
  y = seq(0, 1, length.out = 35)
) |>
  mutate(value = sin(5 * x) + cos(6 * y))

p_toy <- ggplot(raster_grid, aes(x, y, fill = value)) +
  geom_raster() + coord_equal() + labs(title = "A toy raster")

p_thresh <- raster_grid |>
  mutate(high = value > 0.75) |>
  ggplot(aes(x, y, fill = high)) +
  geom_raster() + coord_equal() + labs(title = "value > 0.75")

p_toy + p_thresh
```

## Code

```{r}
#| eval: false
raster_grid <- crossing(
  x = seq(0, 1, length.out = 35),
  y = seq(0, 1, length.out = 35)
) |>
  mutate(value = sin(5 * x) + cos(6 * y))

p_toy <- ggplot(raster_grid, aes(x, y, fill = value)) +
  geom_raster() + coord_equal() + labs(title = "A toy raster")

p_thresh <- raster_grid |>
  mutate(high = value > 0.75) |>
  ggplot(aes(x, y, fill = high)) +
  geom_raster() + coord_equal() + labs(title = "value > 0.75")

p_toy + p_thresh
```
:::

## A real raster: elevation in Luxembourg

`terra` ships with a small example GeoTIFF.

```{r}
elev <- rast(system.file("ex/elev.tif", package = "terra"))
elev
```

Read the printout like a checklist: **dimensions** (rows × columns), **resolution** (size of one cell), **extent** (the bounding box), **coord. ref.** (the CRS), and the value range.

## Plotting a raster with `ggplot2`

One simple route: turn the grid into a data frame with one row per cell.

```{r}
elev_df <- as.data.frame(elev, xy = TRUE)
head(elev_df, 3)
```

```{r}
#| fig-height: 3.5
ggplot(elev_df, aes(x, y, fill = elevation)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_quickmap() +
  labs(fill = "meters")
```

The `tidyterra` package adds `geom_spatraster()` if you'd rather skip this step.

## Resolution is a choice: scale matters

`aggregate()` merges blocks of cells into bigger cells (here, 5 × 5 blocks, averaged).

```{r}
elev_coarse <- aggregate(elev, fact = 5, fun = "mean", na.rm = TRUE)
res(elev)
res(elev_coarse)
```

```{r}
#| echo: false
#| fig-height: 3.5
plot_elev <- function(r, title) {
  as.data.frame(r, xy = TRUE) |>
    ggplot(aes(x, y, fill = elevation)) +
    geom_raster() +
    scale_fill_viridis_c(limits = c(140, 550)) +
    coord_quickmap() +
    labs(title = title, x = NULL, y = NULL, fill = "meters")
}
plot_elev(elev, "original cells") + plot_elev(elev_coarse, "5 × 5 averages") +
  plot_layout(guides = "collect")
```

Coarser cells hide local extremes. The **scale** you analyze at can change your answer.

## Vector meets raster

A classic question: **what is the average raster value inside each polygon?**

::: panel-tabset

## Plot
```{r}
#| echo: false
#| warning: false
lux <- st_read(system.file("ex/lux.shp", package = "terra"), quiet = TRUE)

lux_elev <- terra::extract(elev, vect(lux), fun = mean, na.rm = TRUE)

lux |>
  mutate(mean_elev = lux_elev$elevation) |>
  ggplot() +
  geom_sf(aes(fill = mean_elev), color = "white") +
  scale_fill_viridis_c() +
  labs(title = "Average elevation by canton", fill = "meters")
```

## Code

```{r}
#| eval: false
lux <- st_read(system.file("ex/lux.shp", package = "terra"), quiet = TRUE)

lux_elev <- terra::extract(elev, vect(lux), fun = mean, na.rm = TRUE)

lux |>
  mutate(mean_elev = lux_elev$elevation) |>
  ggplot() +
  geom_sf(aes(fill = mean_elev), color = "white") +
  scale_fill_viridis_c() +
  labs(title = "Average elevation by canton", fill = "meters")
```

:::

`sf` handles the polygons, `terra` handles the grid, and `vect()` passes shapes between them. (Both layers must share a CRS; check first!)

<!-- # Part 3: Spatial statistics, by simulation -->

# Part 3: Spatial correlation

## Why statisticians care about space

Intro statistics usually assumes observations are **independent**. Spatial data often aren't:

> Observations close together in space tend to be more alike than observations far apart.

- Nearby houses have similar prices.
- Nearby temperatures are similar.
- Nearby counties have similar disease rates.

This is sometimes called **Tobler's first law of geography**. It means location carries statistical information, and nearby observations partly **repeat** each other.

<!-- ## Three statistical views of space -->

<!-- | View | Data look like | Example question | -->
<!-- |---|---|---| -->
<!-- | **Point pattern** | where events happened | Are trees clustered? | -->
<!-- | **Areal data** | values attached to regions | Which counties have high rates? | -->
<!-- | **Geostatistical data** | measurements at chosen locations | What is pollution *between* sensors? | -->

<!-- A fourth setting is **trajectory data**: one object moving through space over time. -->

<!-- ::: {.fragment} -->
<!-- **Back to Exercise 1:** coffee shops → point pattern; county unemployment → areal; weather stations → geostatistical; turtle → trajectory; satellite image → a gridded (raster) field. -->
<!-- ::: -->

<!-- ::: {.notes} -->
<!-- Key distinction for point patterns vs geostatistics: in a point pattern the locations themselves are the random outcome; in geostatistics we chose where to measure and the measured value is the outcome. -->
<!-- ::: -->

## Simulation: what does spatial structure look like?

Left: values have **nothing** to do with location. Right: values depend smoothly on location, plus a little noise.

```{r}
#| echo: false
n <- 300
simulation1 <- tibble(x = runif(n), y = runif(n),
                value = rnorm(n))
simulation2 <- tibble(x = runif(n), y = runif(n),
                 value = sin(3 * x) + cos(4 * y) + rnorm(n, sd = 0.2))

map_points <- function(df, title) {
  ggplot(df, aes(x, y, color = value)) +
    geom_point(size = 2.5) + coord_equal() +
    scale_color_viridis_c() + labs(title = title)
}
map_points(simulation1, "simulation1") + map_points(simulation2, "simulation2")
```

## Spatial autocorrelation

**Spatial autocorrelation** = values at nearby locations are related to one another.

::: {.columns}
::: {.column width="50%"}
**Positive** (most common)

- high near high, low near low
- smooth patches
:::
::: {.column width="50%"}
**Negative** (rarer)

- high near low
- checkerboard-like patterns
:::
:::

A useful mental model: each location asks its nearest neighbor, **"are you like me?"**

Let's make that question into a number.

## "Are you like your neighbor?"

For each point, find its **nearest neighbor** (with `sf`!) and record the neighbor's value. Then compute an ordinary correlation.

```{r}
#| label: neighborly
add_neighbor <- function(df, label) {
  nn <- st_nearest_feature(st_as_sf(df, coords = c("x", "y")))  # index of nearest other point
  df |> 
    mutate(neighbor_value = value[nn], data = label)
}

nn_pairs <- bind_rows(add_neighbor(simulation1, "no structure"),
                      add_neighbor(simulation2, "smooth"))

nn_pairs |>
  group_by(data) |>
  summarize(r = cor(value, neighbor_value))
```

<!-- ```{r} -->
<!-- #| fig-height: 3 -->
<!-- ggplot(nn_pairs, aes(value, neighbor_value)) + -->
<!--   geom_point(alpha = 0.4) + facet_wrap(~ data, scales = "free") -->
<!-- ``` -->

<!-- ## Is that correlation real? Shuffle and see. -->

<!-- If location didn't matter, **shuffling the values among the locations** shouldn't change anything. So: shuffle many times and see what correlations chance alone produces. -->

<!-- ```{r} -->
<!-- #| label: cor-real -->
<!-- nn <- st_nearest_feature(st_as_sf(simulation2, coords = c("x", "y"))) -->
<!-- observed <- cor(simulation2$value, simulation2$value[nn]) -->

<!-- shuffled <- replicate(1000, { -->
<!--   v <- sample(simulation2$value)  # same values, random locations -->
<!--   cor(v, v[nn]) -->
<!-- }) -->

<!-- ggplot(tibble(r = shuffled), aes(r)) + -->
<!--   geom_histogram(bins = 30) + -->
<!--   geom_vline(xintercept = observed, color = "red", linewidth = 1.5) + -->
<!--   labs(x = "neighbor correlation", title = "Gray: shuffled worlds.  Red: our data.") -->
<!-- ``` -->

<!-- This is the same **permutation (randomization) test** from intro stats. Real tools like Moran's I (`spdep`) are built on this idea. -->

## Further reading

- *Geocomputation with R* (Lovelace, Nowosad & Muenchow): <https://r.geocompx.org>
- *Spatial Data Science with Applications in R* (Pebesma & Bivand): <https://r-spatial.org/book>
