---
title: "Lab 1: if it's a function, fix it"
subtitle: "Due Tuesday September 8 at 5:00pm"  
editor: source
execute:
  eval: false
---

By the end of the lab, you will...

-   write functions and loops with careful checks and control flow

# Getting started

<!-- -   If you did not complete [lab 0](/labs/lab0.html), then go back and do so to set up your SSH key, git, and find instructions on how to clone a project. -->

-   Next in the *terminal* tab, type `cd ~` to navigate to your home directory. Next `git clone git@github.com:sta323-fa26/lab-1-username.git` where `username` is replaced with your username (this is the copied SSH URL from your lab-1 GitHub repo).

-   Navigate to your `lab-1` folder and open the `.Rproj` file.

-   Open the Quarto (`.qmd`) file, change the author name to your name.

## Exercises

For all exercises, you should respond in the space provided in the template `lab-1.qmd` and show all your work. In all answers of type `double`, three significant figures suffices.

### 1.  **Robust quadratic root finding**

*The goal of this exercise is to avoid numerical instability by re-working the problem analytically.*

The [quadratic formula](https://en.wikipedia.org/wiki/Quadratic_formula) finds $x$ that satisfy the following equation:

$$
ax^2 + bx + c = 0.
$$

In general, $x^*$ such that $f(x^*) = 0$, then $x^*$ is said to be a *root* of $f(x)$. You can find an example of basic implementation of the quadratic formula from [lecture 2: control flow, loops and functions](/slides/lec02.html#/function-composition).

Write a function named `robustQuadraticRoots()` that accepts three arguments: `a`, `b` and `c` and returns both roots (even if they are repeated) while satisfying the additional cases below. Show the output of your function `robustQuadraticRoots()` on each of the cases below.

```{r}
#| eval: false
# case i 
robustQuadraticRoots(a = 0, b = 3, c = 2)

# case ii
robustQuadraticRoots(a = 0, b = 0, c = 0)

# case iii
robustQuadraticRoots(a = 1, b = 2, c = 3)

# case iv
robustQuadraticRoots(a = 4.3162933554e-11,
                       b = 1.0361118521e6,
                       c = -4.5813932360e5)

# case v
robustQuadraticRoots(a = 0, b = 5, c = 0)
```

::: callout-tip
## Hint
- a robust implementation can be derived by multiplying the quadratic formula by "1" in a fancy way.
- for case (ii) you should return an error. Since errors will stop your `.qmd` from rendering, you should include in your code chunk `#| error: true`. Read [quarto docs here](https://quarto.org/docs/computations/execution-options.html) for a full list of code chunk execution options.
- for case (iii) you need to deal with complex numbers [read here](https://www.r-tutor.com/r-introduction/basic-data-types/complex)
- for case (iv) the correct answer is $x_1 = -2.4e16$ and $x_2 = .44$ when rounded to two significant figures.
- for case (v), you should use `warning()` to warn the user that both *a* and *c* are 0.
:::

In each case above, describe why `quadraticRoots()` from the lecture fails.

### 2. **Stirling numbers of the second kind**

*The goal of this exercise is to write a recursive function.* Definition: a **recursive function** is a function that calls itself.

Stirling numbers of the second kind count the number of possible partitions of a set of $n$ objects into $k$ disjoint blocks. For example, the number of ways to partition a set of 3 objects in 2 disjoint blocks is $S(3, 2) = 3$ because the set $\{1, 2, 3\}$ can be partitioned into two disjoint blocks in 3 ways:

$$
\begin{aligned}
&\{1, 2\} \cup \{3\},\\
&\{1, 3\} \cup \{2\},\\
&\{2, 3\} \cup \{1\}
\end{aligned}
$$
We can generate Stirling numbers of the second kind, $S(n,k)$, recursively:

$$
S(n,k) = S(n-1, k-1) + kS(n-1, k)
$$

where we start the sequence with $S(n, 1) = 1$ and the boundary condition $S(n, k) = 0$ for $k>n$.

Create a recursive function `ssk()`that computes **S**tirling numbers of the **s**econd **k**ind. Print the result of your function for the following inputs:

```{r} 
#| eval: false

# case i
ssk(n = 3, k = 2)

# case ii
ssk(n = 10, k = 1)

# case iii
ssk(n = 10, k = 10)

# case iv
ssk(n = 15, k = 8)

# case v
ssk(n = 6, k = 7)
```

### 3. **Secret messages**

#### Introduction

Computers store information using only two values: **0** and **1**.

A single 0 or 1 is called a **bit**. Eight bits grouped together make a **byte**. A byte can represent 256 different values, since

$$
2^8 = 256.
$$

Text, images, and every other type of file on your computer is represented using long sequences of bits. For example, consider the message,

```{r}
message <- "statistics is the best major"
```

R stores each character in the message as a numeric byte. We can look at these bytes using the function

```{r}
charToRaw(message)
```

which converts the characters into their underlying raw byte values. We can go one level deeper, to bits, with

```{r}
message_bits <- rawToBits(charToRaw(message))
message_bits
```

Now we are looking directly at the 0s and 1s used to represent the message. Similarly, we can reverse the process:

```{r}
originalMessage <- rawToChar(packBits(message_bits))
originalMessage
```

`packBits()` groups the bits back into bytes, and `rawToChar()` interprets those bytes as characters again. So conceptually we have gone from

$$
\text{text}
\longrightarrow
\text{bytes}
\longrightarrow
\text{bits}
\longrightarrow
\text{bytes}
\longrightarrow
\text{text}.
$$

This idea is useful because once data have been represented as bits, we can manipulate those bits mathematically.

#### Encrypting with XOR

One simple operation on bits is called **XOR**, pronounced "exclusive OR." For two bits, XOR follows this rule:

| First bit | Second bit | XOR |
| --------- | ---------- | --- |
| 0         | 0          | 0   |
| 0         | 1          | 1   |
| 1         | 0          | 1   |
| 1         | 1          | 0   |

In other words, XOR produces a 1 when the two bits are different and a 0 when they are the same. Suppose part of our message is

$$
1\ 0\ 1\ 1
$$

and we have a secret key

$$
0\ 1\ 1\ 0.
$$

Applying XOR gives

$$
1\ 1\ 0\ 1.
$$

That new sequence can be viewed as encrypted data. The especially useful property of XOR is that applying the **same key again** reverses the operation. If

$$
\text{ciphertext}
=
\text{message}\oplus\text{key},
$$

then

$$
\text{ciphertext}\oplus\text{key}
=
\text{message}.
$$

If you know the key, you can both encrypt and decrypt the message using exactly the same operation. A true **one-time pad** uses a completely random key that is as long as the message and is never reused. Under those conditions, the encryption can provide "perfect" secrecy. For this exercise, however, we will deliberately use a much weaker version so that you can break it.

### Your task

#### Part 1 

Your task is write a function `decrypt()` that takes two inputs `message` and `key` and produces the decrypted message. 

Show your result for the following:

```{r}
secret_message_1 <- c(0,1,1,1,1,0,0,0,0,1,1,1,0,1,0,1,1,0,1,0,1,0,1,
                      1,0,1,1,1,0,0,0,0,0,1,1,1,1,1,0,1,1,0,1,0,0,0,
                      1,1,0,1,1,1,1,0,0,0)
key <- c(1, 0, 1) 

decrypt(message = secret_message_1, key = key)

decrypt(message = secret_message_1, key = c(1,0,2))

decrypt(message = "hi", key = c(1, 0 ,1))
```

Note: the original message was converted to bits and then XORed with a short binary key to create `secret_message`

Importantly, in your final solution:

- Remove the generation of any warnings in your code. Hint: read about the function `rep` using `?rep`.
- Make sure only appropriate keys and messages can be passed in to your function `decrypt()`, i.e. both the key and message must be comprised exclusively of 0s and 1s.


::: callout-tip
## Hint
- the binary key gets vector recycled to match the length of the message
- you can convert from one type to an integer vector by coercion using `as.integer()`, e.g. try  `as.integer(TRUE)`.
:::



#### Part 2

Decrypt the following message, using your function from part 1 and appropriate loops to brute force the solution.


```{r}
secret_message_2 <- c(1,0,1,1,0,0,0,0,0,1,0,0,0,1,0,0,0,0,1,0,0,1,1,0,1,0,1,
                      1,1,1,0,0,1,1,0,1,0,1,0,0,1,0,0,0,1,1,0,0,0,0,1,0,0,1,
                      1,0,0,0,0,0,1,1,0,0,1,1,0,1,0,1,0,0,0,1,1,0,1,1,0,0,0,
                      0,0,0,1,1,0,0,1,0,0,0,1,1,0,0,0,1,1,0,1,1,0,0,1,0,0,0,
                      0,1,0,0,0,0,1,0,0,1,1,0,0,0,0,0,1,1,0,0,0,0,1,1,0,1,0,
                      0,1,0,0,0,0,1,0,0,0,0,1,0,0,1,1,0,0,0,0,0,0,1,0,0,1,0,
                      1,0,0,1,0,0,0,0,0,0,1,1,0,0,1,0,1,0,0,1,0,0,0,0,1,0,0,
                      1,1,0,0,0,0,1,0,1,0,0,1,1,0,1,0,1,0,0,0,1,0,1,0,1,0,0,
                      1,1,0,0,0,1,0,0,0,0,1,0,0,1,1,0,1,0,0,0,0,1,0,0,0,1,0,
                      1,0,1,0,0,1,1,0,1,0,1,0,0,1,0,0,0,1,1,0,0,1,1,0,0,0,1,
                      0,0,0,0,1,1,0,1,0,0,0,0,0,1,0,1,1,0,0,0,1,0,0,1,1,0,1,
                      0,1,1,0,1,0,0,0,0,0,0,1,1,0,0,0,0,1,0,0,1,1,0,1,1,0,0,
                      1,1,0,0,1,0,1,1,0,1,0,0,0,0,0,1,0,1,0,0,0,0,0,1,0,1,0,
                      0,0,0,1,0,0,1,1,0,1,1,1,0,0,1,0,0,1,1,0,1,0,1,0,0,0,1,
                      0,1,0,1,0,0,0,1,0,0,0,1,0,0,1,0,0,0,0,1,0,0,1,1,1,0,1,
                      1,0,0,1,1,1,0,1,1,0,0,0,0,1,0,0,1,1,0,0,0,0,0,1,1,0,0,
                      1,1,0,1,0,1,0,0,0,0,1,0,0,1,1,0,1,0,1,0,0,1,0,0,0,1,0,
                      1,0,1,0,0,1,0,1,1,1,1,0,0,0,0,0,0,1,1,0,0,0,0,1,1,0,1,
                      0,0,1,0,1,1,0,1,0,0,0,1,0,1,0,1,0,0,1,1,0,0,0,1,0,0,0,
                      1,0,1,0,1,1,0)
```


You know only the following:

* the key consists entirely of 0s and 1s,
* the key is short and has length of either 2, 3 or 4,
* the short key was **vector recycled** across the entire message.

::: callout-tip 
## Hint 
- check out the function `intToBin()` in the package `Rutils`, which can be installed with `install.packages("Rutils")` 
:::

## Style guidelines

All assignments in this course must employ proper coding style, as outlined below:

-   All code should obey the 80 character limit per line (i.e. no code should run off the page when rendering or require scrolling). To enable a vertical line in the RStudio IDE that helps guide this, see the style guidelines from lab 0 or ask a member of the teaching team for help.

-   All commas should be followed by a space.

-   All binary operators should be surrounded by space. For example `x + y` is appropriate. `x+y` is not.

-   All pipes `%>%` or `|>` as well as ggplot layers `+` should be followed by a new line.

-   You should be consistent with stylistic choices, e.g. only use 1 of `=` vs `<-` and `%>%` vs `|>`

-   Your name should be at the top (in the YAML) of each document under "author:"

-   All code chunks should be named (with names that don't have spaces, e.g. `ex-1`, `ex-2` etc.)

-   File names in your GitHub repo such as `lab-x.qmd` must not be changed and left as provided. Additionally, your repo must pass certain basic checks. The results of these checks are visible on GitHub via the badges at the top of your README and the actions tab. These are meant to give you feedback around the structure and reproducibility of your repository and assignment - they do not assess the correctness of your work. You should consider them a necessary but not sufficient condition when turning in your work - passing all of the checks simply means your have met a minimum standard of reproducibility for the assignment.

Fundamentally, the check is making sure 1) you only have the files you should in your repository, 2) your `.qmd` renders.

If you have any questions about style, please ask a member of the teaching team.


## Submitting your lab

To submit your assignment, simply commit and push your completed `lab-x.qmd` to your GitHub repo. Your most recent commit 48 hours after the assignment deadline will be graded, and any applicable late penalty will be applied (see [the syllabus](/syllabus.html)). For this reason, **do not push commits** after you are satisfied with your work, or a late penalty will be applied.
