---
title: "Lab 2: walking paths, climbing trees, and finding nests"
subtitle: "Due Monday September 13 at 5:00pm"  
editor: source
execute:
  eval: false
---

By the end of the lab, you will...

- be able to traverse your filesystem recursively in R
- construct, and wrangle nested lists
- work with S3 objects and write S3 methods

## Getting started

-   If you did not complete [lab 0](/labs/lab0.html), then go back and do so to set up your SSH key, git, and find instructions on how to clone a project.

-   Next in the *terminal* tab, type `cd ~` to navigate to your home directory. Next `git clone git@github.com:sta323-fa26/lab-2-username.git` where `username` is replaced with your username (this is the copied SSH URL from your lab-2 GitHub repo).

-   Navigate to your `lab-2` folder and open the `.Rproj` file.

-   Open the Quarto (`.qmd`) file, change the author name to your name.

## Exercises

For all exercises, you should respond in the space provided in the template `lab-2.qmd` and show all your work.

## 1. File systems

*The purpose of this exercise is to become well acquainted with filesystems, lists and recursion*.

#### Exploring a directory recursively

A computer's file system has a naturally recursive structure: a directory can contain files, but it can also contain other directories, which can themselves contain files and directories.

In this exercise, you will write a function

```r
read_tree(path)
```

that converts the directory or file at `path` into a nested R list describing the file structure.

::: callout-tip
## Hint
You may find the following functions helpful: 

- `list.files()`
- `file.info()`
- `basename()`

To start, try running `list.files(".")` to examine the contents of the current directory. 
:::

#### Part 1

Write `read_tree(path)` according to the following rules.

**If `path` refers to a file**, return a named list with four elements:

```r
list(
  name = ...,       # name of the file
  path = ...,       # original path
  type = "file",
  size = ...        # size in megabytes 
)
```

For example, the result might look like

```r
$name
[1] "analysis.R"

$path
[1] "R/analysis.R"

$type
[1] "file"

$size
[1] .004
```

Note: by default, `file.info` will not report megabytes. 

**If `path` refers to a directory**, return a named list containing

```r
list(
  name = ...,
  path = ...,
  type = "directory",
  size = NA_real_,
  children = ...
)
```

Here, `children` should be a list containing the result of `read_tree()` for **every file and directory immediately inside `path`**. So, if a directory contains three items, its `children` element should be a list of length three. Importantly, one of those children might be a directory. In that case, the same procedure must be repeated for the directory. For example, calling `read_tree("/path/to/project")` should return a list of nested lists that looks like:

```text
project
├── README.md
└── data
    ├── raw.csv
    └── clean.csv
```

Your function should ignore hidden ("dot") files. Also, your function should work for directories of **arbitrary depth**. To achieve this, you'll need to write a recursive function (i.e. it will need to call itself, note: this does not mean it must return itself).
Test your function on the directory containing this lab assignment by running the following (with the project open / in the directory of the project root).

```r
tree <- read_tree(".")
str(tree)
```

#### Part 2

Write a new function, `read_tree_2()` that calls `read_tree()` and returns an S3 object of class `fileTree` with the following attributes:

- `root`: the name of the root directory
- `num_files`: the total number of files contained within root and all of its subdirectories
- `total_size`: the total disk space (in megabytes) of all the files within root
- `max_depth`: the maximum depth of a file from root (i.e. if `root` contains `lab-2.qmd` and there are no folders, the depth = 1. If `root` contains 1 subdirectory `data/` which contains `df.csv` and no other directories, the max depth is 2, etc.)
- `largest_file`: the single largest file contained within root or any of is subdirectories

Next, write an S3 `print` function to handle this class. The output of the `print` function, when applied to an object of class `fileTree` should be formatted as follows (actual values will be different): 

```
File tree
---------
Root: lab-1 
Files: 3 
Total size: 0.03 MB
Max depth: 1 


Largest file: lab-1.Rproj
Largest file size: 0.02 MB
```

Report the output of

```{r}
#| eval: false

tree <- read_tree(".")
print(tree)
```

where `"."` is the `lab-2-username` root directory.

::: callout-tip
## Hint

`cat()` is a useful function for printing to the screen. `cat("\nHello")` will print a new line (due to `\n`) followed by "Hello". Additionally, you can mix and match text with objects nicely, e.g.

```
x <- 2
cat("The value of x is ", x)
```
:::

## 2. Earthquakes

*The goal of this exercise is to wrangle raw, moderately sized data*.

We'll revisit the wrangled data set again, later in the semester.

The data set `earthquakes.rds`, loaded via:

```{r}
#| eval: false
earthquakes <- readRDS("data/earthquakes.rds")
```

contains all recorded earthquakes of magnitude greater than, or equal to 2.5 from the last thirty days (data sourced from [U.S. Geological Survey (USGS)](https://earthquake.usgs.gov/earthquakes/feed/v1.0/geojson.php) on September 6, 2026.

The data set is of moderate size and might crash your R instance if you attempt to print the whole thing to the screen. Instead, peek at the structure with careful commands, e.g. 

```{r}
#| eval: false
str(earthquakes, max.level = 1)
str(earthquakes$features[1])
```


<!-- #### Part 1 -->

Produce a "tidy" data frame where each row is an earthquake, and each column describes its properties, id or location (place, longitude, latitude). Next, write a function `remove_redundant()` that deletes any columns from the data frame that either (1) contain all the same value or (2) contain all missing values and prints to the screen which columns are being deleted. Call your final data frame `eq_df` and show the output of the following:

```{r}
#| eval: false
dplyr::glimpse(eq_df)
```

**Questions**: use appropriate subsetting to address the following questions:

1. Which earthquake in the data has the largest magnitude? Report (only) its id, magnitude,
place, longitude, latitude, and depth.
2. What proportion of earthquakes have a non-missing `felt` value?

::: callout-tip
## Hint

- See the [notes](https://sta323-fa26.github.io/slides/lec06.html#/unnest) on `unnest` from the `tidyr` package.

- `which.max` can help with "question" 1
:::


<!-- #### Part 2 -->

<!-- Write a function `missing_report()` that takes a data frame and returns a data frame containing three columns: -->

<!--   - `variable`, name of the variable in the original data frame -->
<!--   - `missing`, the number of `NA` values, and -->
<!--   - `percent_missing`, the percentage of observations that are missing. -->

<!-- Additionally, the rows of the data frame described above should be sorted from greatest to least number of missing observations. -->

<!-- Print the output of  -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- missing_report(eq_df) -->
<!-- ``` -->

<!-- and discuss anything of interest. -->

<!-- **Questions**: The variable `felt` gives the number of responses submitted to the USGS "Did -->
<!-- You Feel It?" system when that information is available. -->

<!-- 1. What proportion of earthquakes have a non-missing `felt` value? -->
<!-- 2. Among earthquakes with a recorded `felt` value, which earthquake received -->
<!--    the most reports? Report its magnitude, place, and number of reports. -->

<!-- In the case of a tie, return all tied earthquakes. -->

<!-- Explain briefly why replacing every missing `felt` value by zero would answer -->
<!-- a different question. -->


## Style guidelines

All assignments in this course must employ proper coding style, as outlined below:

-   All code should obey the 80 character limit per line (i.e. no code should run off the page when rendering or require scrolling). To enable a vertical line in the RStudio IDE that helps guide this, see the style guidelines from lab 0 or ask a member of the teaching team for help.

-   All commas should be followed by a space.

-   All binary operators should be surrounded by space. For example `x + y` is appropriate. `x+y` is not.

-   All pipes `%>%` or `|>` as well as ggplot layers `+` should be followed by a new line.

-   You should be consistent with stylistic choices, e.g. only use 1 of `=` vs `<-` and `%>%` vs `|>`

-   Your name should be at the top (in the YAML) of each document under "author:"

-   All code chunks should be named (with names that don't have spaces, e.g. `ex-1`, `ex-2` etc.)

-   File names in your GitHub repo such as `lab-x.qmd` must not be changed and left as provided. Additionally, your repo must pass certain basic checks. The results of these checks are visible on GitHub via the badges at the top of your README and the actions tab. These are meant to give you feedback around the structure and reproducibility of your repository and assignment - they do not assess the correctness of your work. You should consider them a necessary but not sufficient condition when turning in your work - passing all of the checks simply means your have met a minimum standard of reproducibility for the assignment.

Fundamentally, the check is making sure 1) you only have the files you should in your repository, 2) your `.qmd` renders.

If you have any questions about style, please ask a member of the teaching team.

## Submitting your lab

To submit your assignment, simply commit and push your completed `lab-x.qmd` to your GitHub repo. Your most recent commit 48 hours after the assignment deadline will be graded, and any applicable late penalty will be applied (see [the syllabus](/syllabus.html)). For this reason, **do not push commits** after you are satisfied with your work, or a late penalty will be applied.
