---
title: "Regular expressions cheat sheet"
author: "Dr. Alexander Fisher"
execute:
  warning: true
  eval: true
format: html

---

```{r}
#| message: false
#| warning: false
library(tidyverse)
```

## The basic idea

A **regular expression** (regex or regexp) is a pattern used to find text.
In `stringr`, the string is usually the first argument and the pattern is
the second.

```{r}
#| eval: false
str_detect(strings, pattern)       # Is there a match? TRUE or FALSE
str_subset(strings, pattern)       # Keep strings containing a match
str_count(strings, pattern)        # Count matches in each string
str_locate(strings, pattern)       # Return positions of the first match
str_extract(strings, pattern)      # Return the first match
str_extract_all(strings, pattern)  # Return every match (as a list)
str_remove(strings, pattern)       # Remove the first match
str_remove_all(strings, pattern)   # Remove every match
str_replace(strings, pattern, replacement)
str_replace_all(strings, pattern, replacement)
str_split(strings, pattern)        # Split each string (returns a list)
str_split_1(string, pattern)       # Split one string (returns a vector)
str_match(strings, pattern)        # Return match and captured groups
str_view(strings, pattern)         # Display matches for debugging
```

::: {.callout-tip title="Choose the function from the desired output"}
Use `str_detect()` for a logical vector, `str_subset()` for the original
strings, and `str_extract()` for only the matching text.

Functions ending in `_all()` act on every match within each string. Without
`_all()`, only the first match in each string is used.
:::

## Essential pattern pieces

### Literals, alternatives, and groups

| Goal | Regex | R string | Meaning |
|---|---|---|---|
| Exact text | `cat` | `"cat"` | The characters `cat` |
| One of several characters | `[abc]` | `"[abc]"` | `a`, `b`, or `c` |
| A range | `[A-Z]` | `"[A-Z]"` | One uppercase letter |
| Anything except a set | `[^0-9]` | `"[^0-9]"` | One non-digit |
| Either expression | `cat\|dog` | `"cat|dog"` | `cat` or `dog` |
| Group expressions | `gr(a\|e)y` | `"gr(a|e)y"` | `gray` or `grey` |
| Group without capture | `(?:cat\|dog)` | `"(?:cat|dog)"` | Group only |

```{r}
#| label: gray
str_detect(c("gray", "grey"), "gr(a|e)y")

names <- c("aang", "katara", "sokka")

# A vowel at the start, not a vowel anywhere or "a" at the start.
str_detect(names, "^[aeiou]")
```

### Character classes

- Notice we need to double escape in R because the backslash is already an escape character.

| Regex | R string | Matches one... |
|---|---|---|
| `.` | `"."` | Character other than a newline |
| `\d` | `"\\d"` | Digit |
| `\D` | `"\\D"` | Non-digit |
| `\s` | `"\\s"` | Whitespace character |
| `\S` | `"\\S"` | Non-whitespace character |
| `\w` | `"\\w"` | Letter, digit, or underscore |
| `\W` | `"\\W"` | Non-word character |
| `[[:alpha:]]` | `"[[:alpha:]]"` | Alphabetic character |
| `[[:digit:]]` | `"[[:digit:]]"` | Digit |
| `[[:punct:]]` | `"[[:punct:]]"` | Punctuation character |

The dot is a wildcard. To match a literal dot, use `\\.` in an ordinary R
string.

```{r}
str_detect("duke.edu", ".") # matching a wildcard
str_detect("duke.edu", "\\.")     
str_detect("dukeXedu", "\\.")     
```

### Anchors and boundaries

Anchors match a **position**, not a character.

| Regex | R string | Position |
|---|---|---|
| `^` | `"^"` | Start of string |
| `$` | `"$"` | End of string |
| `\b` | `"\\b"` | Word boundary |
| `\B` | `"\\B"` | Not a word boundary |

```{r}
#| eval: false
str_detect(x, "^From:")           # starts with From:
str_detect(x, "\\.edu$")          # ends with .edu
str_detect(x, "\\bcat\\b")        # the whole word cat
str_detect(x, "^\\s*$")           # blank or whitespace-only string
```

```{r}
#| label: example-read-lines

readLines("https://sta323-fa26.github.io/data/emails.txt")
```


::: {.callout-important title="A blank line read by readLines()"}
`readLines()` returns one character string per line. A blank line is therefore
usually an element equal to `""`; it is not stored as `"\n"`. The pattern
`"^\\s*$"` matches both `""` and a line containing only whitespace.
:::


### Exercise 1

Extract the emails from above and create a data frame with full email in the first column.

Bonus: extract domain (e.g. "gmail") in the second column (a look around may make this easy, see the end of doc).

### Quantifiers

A quantifier controls how many times the **preceding item** may repeat.

| Quantifier | Meaning | Example |
|---|---|---|
| `*` | Zero or more | `"a*"` |
| `+` | One or more | `"a+"` |
| `?` | Zero or one | `"colou?r"` |
| `{3}` | Exactly three | `"\\d{3}"` |
| `{3,}` | Three or more | `"\\d{3,}"` |
| `{3,5}` | Three through five | `"\\d{3,5}"` |

```{r}
str_extract("919-123-4567", "^\\d{3}-\\d{3}-\\d{4}$")
str_detect(c("color", "colour"), "^colou?r$")
```

Common whitespace templates are:

```{r}
#| eval: false
"\\s*"  # allow no whitespace or any amount of whitespace
"\\s+"  # require at least one whitespace character
"\\s?"  # allow at most one whitespace character
```

### Greedy and lazy matching

Quantifiers are **greedy** by default: they match as much text as possible.
Add `?` after a quantifier to make it lazy (or non-greedy).

```{r}
html <- "<div>first</div><div>second</div>"

str_extract(html, "<div>.*</div>")
# <div>first</div><div>second</div>

str_extract(html, "<div>.*?</div>")
# <div>first</div>
```

## R has two layers of escaping

Characters such as `. ^ $ * + ? { } [ ] \\ | ( )` have special regex
meanings. A backslash tells the regex engine to treat one literally, but the
backslash is also special inside an ordinary R string. Consequently, an
ordinary R string generally needs **two backslashes**.

| Desired match | Regex engine needs | Ordinary R string | Raw R string |
|---|---|---|---|
| A digit | `\d` | `"\\d"` | `r"(\d)"` |
| A period | `\.` | `"\\."` | `r"(\.)"` |
| `(` | `\(` | `"\\("` | `r"(\()"` |
| A backslash | `\\` | `"\\\\"` | `r"(\\)"` |

Raw strings, available in R 4.0 and later, can make complicated patterns much
easier to read.

```{r}
ordinary <- "\\(\\d{3}\\)\\s*\\d{3}-\\d{4}"
raw <- r"(\(\d{3}\)\s*\d{3}-\d{4})"

identical(ordinary, raw)
```

## Pattern options

Wrap a pattern in `regex()` to change how it is interpreted.

```{r}
#| eval: false
# Ignore capitalization.
str_detect(x, regex("duke", ignore_case = TRUE))

# Let ^ and $ refer to the start and end of each line in one string.
str_extract_all(
  one_multiline_string,
  regex("^From:.*$", multiline = TRUE)
)

# Let the . wildcard match newline characters too.
str_extract(
  one_multiline_string,
  regex("BEGIN.*END", dotall = TRUE)
)

# Permit spacing and comments inside a long pattern.
phone_pattern <- regex(
  r"(
    \(? (\d{3}) \)?  # optional parentheses around area code
    [ -]?             # optional separator
    (\d{3})           # exchange
    [ -]?             # optional separator
    (\d{4})           # subscriber number
  )",
  comments = TRUE
)
```

Use `fixed()` when the pattern should have **no regex meaning**. This is often
the clearest and fastest choice for literal text.

```{r}
str_detect("price is $5.00", fixed("$5.00"))
```

## Capture groups and backreferences

Parentheses can save parts of a match. `str_match()` returns the full match in
its first column and one captured group in each additional column.

```{r}
dates <- c("2026-09-15", "2025-01-30")

str_match(dates, "^(\\d{4})-(\\d{2})-(\\d{2})$")
#      full match   year   month  day
```

In a replacement, `\\1`, `\\2`, and so on refer to captured groups.

```{r}
str_replace(
  "NC, Durham",
  "^([^,]+),\\s*(.+)$",
  "\\2 \\1"
)
```

A backreference can also require text to repeat within the match.

```{r}
str_detect(c("abab", "cdcd", "abcd"), "^(..)\\1$")
# (..) matches any two characters in this group
# \1 matches the same as group 1 
# $ signifies end of string
```

## Lookarounds

Lookarounds require context without including that context in the returned
match.

| Form | Meaning |
|---|---|
| `(?=...)` | Must be followed by... |
| `(?!...)` | Must not be followed by... |
| `(?<=...)` | Must be preceded by... |
| `(?<!...)` | Must not be preceded by... |

```{r}
#| eval: false
# Extract digits only when they follow "DUID: ".
str_extract(student_id, "(?<=DUID: )\\d+")

# Extract a username only when it is followed by @.
str_extract(address, "[A-Za-z0-9._+-]+(?=@)")

# Extract a number only when it is not followed by a unit such as kg.
str_extract(x, "\\d+\\b(?!\\s*kg\\b)")
```

::: {.callout-note}
Lookbehind is useful when the prefix has a fixed length. If the surrounding
text is complicated, a capture group with `str_match()` is often clearer.
:::

## Common templates

### Match a complete string

Use both anchors when the **entire** string must follow the format.

```{r}
#| eval: false
str_detect(x, "^[A-Za-z]+$") # any character repeating
```

### Extract words or numbers

```{r}
#| eval: false
str_extract_all(x, "\\b[[:alpha:]]+\\b")
```

The number pattern above allows an optional minus sign and optional decimal
part. The currency pattern allows commas and an optional two-digit cents part.

### Match a file extension

```{r}
#| eval: false
str_detect(
  filenames,
  regex("\\.(csv|tsv|txt)$", ignore_case = TRUE)
)
```

The dot must be escaped, and `$` prevents a match such as `data.csv.backup`.


### Split around a delimiter and optional whitespace

```{r}
str_split_1("rec.autos, rec.autos.tech", "\\s*,\\s*")
str_split_1("key : value", "\\s*:\\s*")
```

### Remove a labeled prefix

```{r}
str_remove("From: name@duke.edu", "^From\\s*:\\s*")
```

Anchoring the label with `^` prevents removal of the word `From` later in the
text.

### Exercise 2

```{r}
#| eval: false
download.file("https://sta323-fa25.github.io/data/secret-message.txt", 
              destfile = "secret-message.txt")
download.file("https://sta323-fa26.github.io/data/secret-messages.zip", 
              destfile = "secret-messages.zip")

unzip("secret-messages.zip", exdir = "secret-messages")
```

In `secret-message.txt`, (and the folder) find the secret message. It will be of the form `sta323{secret-message}` where `secret-message` is replaced by some other text.

::: callout-warning
## WARNING
Do NOT view the files --  RStudio may CRASH!
:::


## Base R equivalents

| Task | `stringr` | Base R |
|---|---|---|
| Logical detection | `str_detect(x, p)` | `grepl(p, x)` |
| Indices of matches | `which(str_detect(x, p))` | `grep(p, x)` |
| Matching strings | `str_subset(x, p)` | `grep(p, x, value = TRUE)` |
| Replace first | `str_replace(x, p, r)` | `sub(p, r, x)` |
| Replace all | `str_replace_all(x, p, r)` | `gsub(p, r, x)` |
| Split | `str_split(x, p)` | `strsplit(x, p)` |

Note the argument order: `stringr` generally uses `(string, pattern)`, whereas
base R's regex functions generally use `(pattern, string)`.

## Debugging checklist

When a pattern does not behave as expected, check these in order:

1. **Inspect the actual strings.** Use `str_view(x)` to expose
   whitespace and escapes.
2. **Start with a small literal.** Confirm that the relevant line can be found
   before adding classes, groups, and quantifiers.
3. **Check anchors.** Should the match occur anywhere, or only at the start or
   end?
4. **Check escaping.** In an ordinary R string, `\\d`, `\\s`, and `\\.` need
   two backslashes.
5. **Check capitalization.** Use `regex(..., ignore_case = TRUE)` when case
   should not matter.
6. **Check missing values.** Most `stringr` functions return `NA` for an `NA`
   input.
7. **Check greediness.** Replace `.*` with a more specific class or use `.*?`
   when the shortest match is intended.


## Further references

- [`stringr` function reference](https://stringr.tidyverse.org/reference/)
